---
title: "Hands-0n Exercise 5-D:"
subtitle:  "Visual Multivariate Analysis with Parallel Coordinates Plot"
date: "12 May 2026"
date-modified: "last-modified"
editor: visual
execute:
  freeze: auto
  echo: true
  eval: true
  warning: false
---

# 1 Overview

Parallel coordinates plot is a data visualisation specially designed for visualising and analysing multivariate, numerical data. It is ideal for comparing multiple quantitative variables together and seeing the relationships between them. Its strength is that the variables can even be completely different: different ranges and even different units.

By the end of this hands-on exercise, I will gain hands-on experience on:

-   plotting statistic parallel coordinates plots by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://r4va.netlify.app/chap15) package, and

-   plotting interactive parallel coordinates plots by using [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) package.

# 2 Getting Started

::: panel-tabset
## Installing & Launching R Packages

For this exercise, the **GGally**, **parcoords**, **parallelPlot** and **tidyverse** packages will be used.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## Importing Data

In this hands-on exercise, the data of [World Happiness 2018 report](https://worldhappiness.report/ed/2018/) will be used. The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

The output is a tibble data frame called **wh**, with 156 observations (rows) across 12 variables (columns).

```{r}
glimpse(wh)
```

```{r}
n_distinct(wh$Country)
```
:::

# 3 Static Parallel Coordinates Plot

Before getting started, it is a good practice to read the function description in detail.

## 3.1 Plotting a simple parallel coordinates

The code chunk below plots a basic parallel coordinate plot.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggparcoord(data = wh, 
           columns = c(7:12)) +
  labs(title = "Basic Parallel Plot") +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(hjust=1, angle=0)
  ) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

## 3.2 Plotting a parallel coordinates with points

`showPoints = TRUE` and `alphaLines = 0.3` are added to show individual observations and reduce opacity of the lines to reduce visual clutter.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggparcoord(data = wh, 
           columns = c(7:12),
           showPoints = TRUE,
           alphaLines = 0.3) +
  labs(title = "Basic Parallel Plot with Points") +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(hjust=1, angle=0)
  ) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

## 3.3 Plotting a parallel coordinates with boxplot

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

Things to learn from the code chunk above.

-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

-   `alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

-   `title` argument is used to provide the parallel coordinates plot a title.

## 3.4 Parallel coordinates with facet

Since `ggparcoord()` is developed by extending **ggplot2** package, we can combination use some of the ggplot2 function when plotting a parallel coordinates plot.

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

## 3.5 Parallel coordinates with palette

**Viridis Color Palette** `groupColumn = "Region"` groups and colours the observations by their Region. We apply `scale_color_viridis_d()` to ensure the color palette is accessible and visually distinct.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggparcoord(data = wh, 
           columns = c(7:12),
           groupColumn = "Region") +
  labs(title = "Basic Parallel Plot, grouped by Region") +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(hjust=1, angle=0),
      legend.text = element_text(size=8),
      legend.position = "bottom"
  ) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

The highlighting effect is performed manually by assigning a highlight colour via `scale_color_manual()` to the group of interest, while muting the colours of the other groups in the backdrop.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggparcoord(data = wh, 
           columns = c(7:12),
           groupColumn = "Region") +
  labs(title = "Basic Parallel Plot, highlighting a group") +
  scale_color_manual(values=c("#E8E8E8", "#E8E8E8", "#E8E8E8", "#E8E8E8", 
                              "#E8E8E8", "#E8E8E8", "#E8E8E8", "#E8E8E8", 
                              "#E8E8E8", "#69b3a2")) +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(hjust=1, angle=0),
      legend.text = element_text(size=8),
      legend.position = "bottom"
  ) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

## 3.6 Parallel Coordinates with Boxplot (2)

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables") +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(hjust=1, angle=0),
      legend.text = element_text(size=8)
  ) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

::: callout-note
-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.
-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scales each variable so the minimum is zero and the maximum is one.
-   `alphaLines` argument is used to reduce the intensity of the line colour to 0.2.
-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`.
:::

## 3.7 Parallel coordinates with facet (2)

Since `ggparcoord()` is developed by extending **ggplot2** package, we can combine it with some ggplot2 functions when plotting.

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represents one geographical region.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 8
#| fig-width: 10

ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots by Region") +
  facet_wrap(~ Region, labeller = labeller(group = label_wrap_gen(width = 25))) +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(
      plot.title = element_text(hjust=0, face = "bold"),
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle=30, hjust=1, size= 8),
      axis.title.y = element_blank(),
      strip.text = element_text(size = 9, face = "bold"),
      legend.position = "none" # Legend is redundant due to faceting
  )
```

## 3.8 Rotating x-axis text label

To make the x-axis text label easy to read, let us rotate the labels by 30 degrees. We can rotate axis text labels using `theme()` function in ggplot2 as shown in the code chunk below

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

Thing to learn from the code chunk above:

-   To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. And we specify `element_text(angle = 30)` to rotate the x-axis text by an angle 30 degree.

## 3.9 Adjusting the rotated x-axis text label

Rotating x-axis text labels to 30 degrees makes the label overlap with the plot and we can avoid this by adjusting the text location using `hjust` argument to theme’s text element with `element_text()`. We use `axis.text.x` as we want to change the look of x-axis text.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

# 4 Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and [d3.js](https://d3js.org/). In this section, you will learn how to use functions provided in parallelPlot package to build interactive parallel coordinates plot.

**Robust Subset and Responsive Sizing** We clean up the `select()` statement to comply with modern `tidyselect` standards, and we remove the hardcoded dimensions so the plot scales responsively to the user's screen.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Robustly select the target variables
wh_interactive <- wh %>%
  select(`Happiness score`, 7:12)

# Generate the interactive plot with continuous color scaling
parallelPlot(wh_interactive,
             rotateTitle = TRUE,
             continuousCS = "YlOrRd")
```

**Interactivity:**

-   **Click** on a variable of interest (e.g., Happiness score). The monotonous blue color will dynamically change to a heatmap-style gradient reflecting the values of that specific column.
-   **Hover** over an observation to highlight its trajectory across all variables.
-   **Click and drag** vertically along any axis to "brush" the data, filtering the visualization to only show countries within that specific range.

## 4.2 Rotate axis label

In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

One of the useful interactive feature of parallelPlot is we can click on a variable of interest, for example Happiness score, the monotonous blue colour (default) will change a blues with different intensity colour scheme will be used.

## 4.3 Changing the colour scheme

We can change the default blue colour scheme by using `continousCS` argument as shown in the code chunl below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## 4.4 Parallel coordinates plot with histogram

In the code chunk below, `histoVisibility` argument is used to plot histograms along the axis of each variable, instantly revealing the underlying distribution of the data.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

histoVisibility <- rep(TRUE, ncol(wh_interactive))

parallelPlot(wh_interactive,
             rotateTitle = TRUE,
             continuousCS = "YlOrRd",
             histoVisibility = histoVisibility)
```

**Interactivity:**

-   Click on a variable of interest, for example Happiness score, the monotonous blue colour (default) will change a blues with different intensity colour scheme will be used.
-   Hover over observation of interest to highlight its trajectory across variables.

## 4.5 Parallel coordinates plot with histogram (2)

In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```
