---
title: "Hands-0n Exercise 5-C:"
subtitle:  "Heatmap for Visualising and Analysing Multivariate Data"
date: "12 May 2026"
date-modified: "last-modified"
editor: visual
execute:
  freeze: auto
  echo: true
  eval: true
  warning: false
---

# 1 Overview

Heatmaps visualise data through variations in colouring. When applied to a tabular format, heatmaps are useful for cross-examining multivariate data, through placing variables in the columns and observation (or records) in rows and colouring the cells within the table. Heatmaps are good for showing variance across multiple variables, revealing any patterns, displaying whether any variables are similar to each other, and for detecting if any correlations exist in-between them.

In this hands-on exercise, we will gain hands-on experience on using R to plot static and interactive heatmap for visualising and analysing multivariate data.

# 2 Getting Started

::: panel-tabset
## Installing & Launching R Packages

```{r}
pacman::p_load(seriation, dendextend, heatmaply, 
               tidyverse, ggridges, ggtext,
               patchwork)
```

## Importing Data

In this hands-on exercise, the data of [World Happiness 2018 report](https://worldhappiness.report/ed/2018/) will be used. The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

```{r}
glimpse(wh)
```
:::

Ultimately, we want a heatmap where the different countries are shown along one axis, the explanatory variables on another axis, and the shading/intensity of the cells should reflect the magnitude of values.

## Understanding the Variables

### Target Variable (Happiness Score)

```{r}
# Calculate Median and Mean values for plotting normal curve
med_hs <- median(wh$`Happiness score`)
mean_hs <- mean(wh$`Happiness score`)
std_hs <- sd(wh$`Happiness score`)

ggplot(
    wh, 
    aes(x = `Happiness score`)
  ) +
  geom_density(
    color = "#d18a7d", fill = "#d18a7d",
    adjust = .2, alpha = .6
  ) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean_hs, sd = std_hs),
    col = "#65666a", linewidth = .6
  ) +
  geom_vline(aes(xintercept = mean_hs),
                 color = "#dc7068", linetype = "dotted", linewidth = .8) +
    annotate("text", x = mean_hs-0.3, y = 0.45,
      label = paste("Mean: ", round(mean_hs, 2)), color = "#dc7068", size = 3.5) +
  geom_vline(aes(xintercept = med_hs),
                 color = "steelblue", linetype = "dotted", linewidth = .8) +  
    annotate("text", x = med_hs+0.4, y = 0.45,
      label = paste("Median: ", round(med_hs, 2)), color = "steelblue", size = 3.5) +  
  labs(
    title = "Uneven Distribution of Happiness Scores",
    subtitle = "Median and Mean happiness indicators are similar, with various humps noted",
    x = "Happiness Score"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    plot.title = element_text(face="bold"),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

### Dependent Variables

**Categorical**

```{r}
ggplot(data = wh, 
       aes(x = fct_rev(fct_infreq(Region)))) +
    geom_bar(fill = "steelblue", alpha = 0.8)+
    ylim(0, 45) + 
    geom_text(stat="count", 
      aes(label=paste0(after_stat(count), ", ", 
      round(after_stat(count)/sum(after_stat(count))*100, 1), "%")),
      vjust= 0.5, hjust = -0.2, size= 3.5) +
    labs(x = "Region",
         y = "No. of Countries",
         title = "Number of Countries by Region") + 
    theme_minimal() +
    theme(plot.title = element_text(face="bold")) +
  coord_flip()
```

**Numerical Distributions**

**Streamlined EDA with facet_wrap** Instead of manually repeating `ggplot` code 8 times for every single variable, we can use `pivot_longer()` to reshape the data and `facet_wrap()` to plot them all instantly. This is a much cleaner and robust "R-way" of handling multiple distributions.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-width: 10
#| fig-height: 8

wh %>%
  select(`Whisker-high`, `Whisker-low`, Dystopia, `GDP per capita`, 
         `Social support`, `Healthy life expectancy`, 
         `Freedom to make life choices`, Generosity, `Perceptions of corruption`) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_density(color = "#EABEC3", fill = "#EABEC3", adjust = 0.2, alpha = 0.6) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(title = "Uneven Distribution across explanatory variables", x = NULL)
```

::: callout-note
**Insights**

Range of each variable is different. Whisker-high/low varies between 2 to 8, while variables like generosity and life expectancy range from 0 to 1.
:::

## Data Preparation

**Modification 2: Robust Matrix Conversion** Instead of forcing row names onto a tibble and using confusing column indices like `wh_matrix[, -c(1, 2, 4, 5)]`, we explicitly select the variables we want and use `column_to_rownames()` to generate a clean, named matrix.

```{r}
# Extract the Region mapping for row annotations later
wh_region <- wh %>% select(Country, Region) %>% column_to_rownames("Country")

# Convert the data frame into a matrix strictly containing numerical indicators
wh_matrix <- wh %>%
  select(Country, `Happiness score`, `GDP per capita`, `Social support`, 
         `Healthy life expectancy`, `Freedom to make life choices`, 
         Generosity, `Perceptions of corruption`, Dystopia) %>%
  column_to_rownames("Country") %>%
  as.matrix()
```

# 3 Static Heatmap

There are many R packages and functions can be used to drawing static heatmaps, they are:

-   [heatmap()](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/heatmap) of R stats package for simple heatmaps.

-   [heatmap.2()](https://www.rdocumentation.org/packages/gplots/versions/3.0.1.1/topics/heatmap.2) of **gplots** R package for enhanced heatmaps.

-   [pheatmap()](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12/topics/pheatmap) of [**pheatmap**](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12) R package.

## 3.1 heatmap() of R Stats

### Basic Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

::: callout-note
By default, `heatmap()` plots a cluster heatmap. The arguments `Rowv=NA` and `Colv=NA` are used to switch off the option of plotting the row and column dendrograms.
:::

### Cluster Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

::: callout-note
-   This heatmap is hard to interpret since we are unable to see colours between the opposing ends of the range of the values.

-   Happiness Score variable have relatively higher values, which makes that the other variables with small values all look the same. Thus, we need to normalize this matrix to "scale" our data.
:::

### Normalised Heatmap

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

::: callout-note
-   Notice that the values are scaled now.

-   `Margins` argument is used to ensure that the entire x-axis labels are displayed completely

-   `cexRow` and `cexCol` arguments are used to define the font size used for y-axis and x-axis labels respectively.
:::

# 4 Interactive Heatmap

[**heatmaply**](http://talgalili.github.io/heatmaply/) is an R package for building interactive cluster heatmap that can be shared online as a stand-alone HTML file.

## 4.1 Working with **heatmaply**

The code chunk below shows the basic syntax needed to create an interactive heatmap by using **heatmaply** package. Because we cleaned our matrix earlier, we simply pass `wh_matrix`.

```{r}
heatmaply(wh_matrix, 
          fontsize_col = 7,
          fontsize_row = 4,
          margins = c(60,130,10,10))
```

```{r}
heatmaply(mtcars)
```

The code chunk below shows the basic syntax needed to create n interactive heatmap by using **heatmaply** package.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

Note that:

-   Different from *heatmap()*, for *heatmaply()* the default horizontal dendrogram is placed on the left hand side of the heatmap.

-   The text label of each raw, on the other hand, is placed on the right hand side of the heat map.

-   When the x-axis marker labels are too long, they will be rotated by 135 degree from the north.

## 4.2 Data transformation

Three main data transformation methods are supported by *heatmaply()*, namely: `scale`, `normalize`, and `percentize`.

### 4.2.1 Scaling method

-   When all variables are came from or assumed to come from some normal distribution, then scaling (i.e.: subtract the mean and divide by the standard deviation) would bring them all close to the standard normal distribution.

-   In such a case, each value would reflect the distance from the mean in units of standard deviation.

-   The *scale* argument in *heatmaply()* supports column and row scaling.

The code chunk below is used to scale variable values columewise

```{r}
heatmaply(wh_matrix,
          fontsize_col = 7,
          fontsize_row = 4,          
          scale = "column")
```

### 4.2.2 Normalising method

-   When variables in the data comes from possibly different (and non-normal) distributions, the normalize function can be used to bring data to the 0 to 1 scale by subtracting the minimum and dividing by the maximum of all observations.

-   This preserves the shape of each variable’s distribution while making them easily comparable on the same “scale”.

Different from Scaling, the normalise method is performed on the input data set i.e. wh_matrix as shown in the code chunk below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

Alternative:

```{r}
heatmaply(normalize(wh_matrix),
          fontsize_col = 7,
          fontsize_row = 4)
```

### 4.2.3 Percentising method

-   This is similar to ranking the variables, but instead of keeping the rank values, divide them by the maximal rank.

-   This is done by using the ecdf of the variables on their own values, bringing each value to its empirical percentile.

-   The benefit of the percentize function is that each value has a relatively clear interpretation, it is the percent of observations that got that value or below it.

Similar to Normalize method, the Percentize method is also performed on the input data set i.e. wh_matrix as shown in the code chunk below.

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

Alternative:

```{r}
heatmaply(percentize(wh_matrix),
          fontsize_col = 7,
          fontsize_row = 4)
```

## 4.3 Clustering algorithm

**heatmaply** supports a variety of hierarchical clustering algorithm. The main arguments provided are:

-   *distfun*: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to dist. The options “pearson”, “spearman” and “kendall” can be used to use correlation-based clustering, which uses as.dist(1 - cor(t(x))) as the distance metric (using the specified correlation method).

-   *hclustfun*: function used to compute the hierarchical clustering when *Rowv* or *Colv* are not dendrograms. Defaults to *hclust*.

-   *dist_method* default is NULL, which results in “euclidean” to be used. It can accept alternative character strings indicating the method to be passed to distfun. By default *distfun* is “dist”” hence this can be one of “euclidean”, “maximum”, “manhattan”, “canberra”, “binary” or “minkowski”.

-   *hclust_method* default is NULL, which results in “complete” method to be used. It can accept alternative character strings indicating the method to be passed to *hclustfun*. By default hclustfun is hclust hence this can be one of “ward.D”, “ward.D2”, “single”, “complete”, “average” (= UPGMA), “mcquitty” (= WPGMA), “median” (= WPGMC) or “centroid” (= UPGMC).

In general, a clustering model can be calibrated either manually or statistically.

### 4.3.1 Manual approach

In the code chunk below, the heatmap is plotted by using hierachical clustering algorithm with “Euclidean distance” and “ward.D” method.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

Alternative:

```{r}
heatmaply(normalize(wh_matrix),
          dist_method = "euclidean",
          hclust_method = "ward.D",
          fontsize_col = 7,
          fontsize_row = 4)

```

### 4.3.2 Statistical approach

We can use `dend_expend()` to determine the recommended clustering method, and `find_k()` to determine the optimal number of clusters.

```{r}
wh_d <- dist(normalize(wh_matrix), method = "euclidean")  

# Find best clustering method
dend_expend(wh_d)[[3]]
```

```{r}
# Find optimum k clusters
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

Figure above shows that **k=3** would be optimal.

```{r}
heatmaply(normalize(wh_matrix),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3,
          fontsize_col = 7,
          fontsize_row = 4
          )
```

## 4.4 Seriation

One of the problems with hierarchical clustering is that it doesn’t actually place the rows in a definite order, it merely constrains the space of possible orderings. Take three items A, B and C. If you ignore reflections, there are three possible orderings: ABC, ACB, BAC. If clustering them gives you ((A+B)+C) as a tree, you know that C can’t end up between A and B, but it doesn’t tell you which way to flip the A+B cluster. It doesn’t tell you if the ABC ordering will lead to a clearer-looking heatmap than the BAC ordering.

**heatmaply** uses the seriation package to find an optimal ordering of rows and columns. Optimal means to optimize the Hamiltonian path length that is restricted by the dendrogram structure. This, in other words, means to rotate the branches so that the sum of distances between each adjacent leaf (label) will be minimized. This is related to a restricted version of the travelling salesman problem.

Here we meet our first seriation algorithm: Optimal Leaf Ordering (OLO). This algorithm starts with the output of an agglomerative clustering algorithm and produces a unique ordering, one that flips the various branches of the dendrogram around so as to minimize the sum of dissimilarities between adjacent leaves. Here is the result of applying Optimal Leaf Ordering to the same clustering result as the heatmap above.

### OLO

```{r}
heatmaply(normalize(wh_matrix),
          seriate = "OLO",
          fontsize_col = 7,
          fontsize_row = 4)
```

The default options is “OLO” (Optimal leaf ordering) which optimizes the above criterion (in O(n\^4)). Another option is “GW” (Gruvaeus and Wainer) which aims for the same goal but uses a potentially faster heuristic.

### GW

```{r}
heatmaply(normalize(wh_matrix),
          seriate = "GW",
          fontsize_col = 7,
          fontsize_row = 4)
```

The option “mean” gives the output we would get by default from heatmap functions in other packages such as gplots::heatmap.2.

### Mean

```{r}
heatmaply(normalize(wh_matrix),
          seriate = "mean",
          fontsize_col = 7,
          fontsize_row = 4
          )
```

The option “none” gives us the dendrograms without any rotation that is based on the data matrix.

### None

```{r}
heatmaply(normalize(wh_matrix),
          seriate = "none",
          fontsize_col = 7,
          fontsize_row = 4
          )
```

## 4.5 **Working with colour palettes**

**Adding Geographic Context with `row_side_colors`** By injecting our previously saved `wh_region` dataframe into the `row_side_colors` parameter, `heatmaply` will append a colored bar along the left axis. This allows us to instantly visualize if certain clusters of variables correlate with specific geographic regions (e.g., are European countries clustering together?).

The default colour palette uses by **heatmaply** is *viridis*. heatmaply users, however, can use other colour palettes in order to improve the aestheticness and visual friendliness of the heatmap.

In the code chunk below, the Blues colour palette of rColorBrewer is used

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

Better usage of palettes:

```{r}
heatmaply(normalize(wh_matrix),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3, 
          seriate = "OLO",
          colors = Blues,
          row_side_colors = wh_region, #<<< Geographic context added here!
          fontsize_col = 7,
          fontsize_row = 4
          )
```

## 4.6 **The finishing touch**

Beside providing a wide collection of arguments for meeting the statistical analysis needs, *heatmaply* also provides many plotting features to ensure cartographic quality heatmap can be produced.

In the code chunk below the following arguments are used:

-   *k_row* is used to produce 5 groups.

-   *margins* is used to change the top margin to 60 and row margin to 200.

-   *fontsizw_row* and *fontsize_col* are used to change the font size for row and column labels to 4.

-   *main* is used to write the main title of the plot.

-   *xlab* and *ylab* are used to write the x-axis and y-axis labels respectively.

```{r}
#| fig-width: 12
heatmaply(normalize(wh_matrix),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```

# 5 Interactive Correlation Heatmap

**Multivariate Correlation Heatmap** Heatmaps aren't just for clustering records (countries)—they are arguably even more useful for clustering *variables* to see how they correlate with one another. Using `heatmaply_cor()`, we can generate an interactive correlation matrix of the World Happiness indicators.

```{r}
#| fig-width: 8
#| fig-height: 8

# Calculate the correlation matrix
wh_cor <- cor(wh_matrix)

# Plot interactive correlation heatmap
heatmaply_cor(
  wh_cor,
  k_col = 2, 
  k_row = 2,
  colors = colorRampPalette(c("#d5695d", "white", "#0072B2"))(100),
  margins = c(50, 50, 50, 50),
  main = "Interactive Correlation Matrix of World Happiness Variables"
)
```
