---
title: "Hands-0n Exercise 5-A:"
subtitle:  "Building Ternary Plot with R"
date: "12 May 2026"
date-modified: "last-modified"
editor: visual
execute:
  freeze: auto
  echo: true
  eval: true
  warning: false
---

# 1 Overview

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It's display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plots programmatically using R for visualising and analysing population structure of Singapore.

# 2 Getting Started

::: panel-tabset
## Installing & Launching R Packages

For this exercise, several R packages will be used:

1.  [**ggtern**](http://www.ggtern.com/d/2.2.0/), a ggplot extension specially designed to plot ternary diagrams.
2.  **Plotly R**, an R package for creating interactive web-based graphs.
3.  **tidyverse** for robust data wrangling.

```{r}
pacman::p_load(plotly, tidyverse, ggtern,
               tricolore, gganimate)
```

## Importing Data

The original data is obtained from [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data).

```{r}
#Reading the data into R environment
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")
```

## Preparing Data

**Robust Data Wrangling using Modern `tidyr`** Instead of using the deprecated `spread()` function and hardcoded indices, we will use modern `pivot_wider()` and select columns explicitly. This ensures the code won't break if the dataset column order changes.

```{r}
# Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year)) %>%
  pivot_wider(names_from = AG, values_from = Population) %>%
  mutate(YOUNG = rowSums(across(`0_to_4`:`20_to_24`)),      # Ages 0-24
         ACTIVE = rowSums(across(`25_to_29`:`60_to_64`)),   # Ages 25-64
         OLD = rowSums(across(`65_to_69`:`90_and_over`)),   # Above 65
         TOTAL = rowSums(across(YOUNG:OLD))) %>%
  filter(Year == 2018) %>%
  filter(TOTAL > 0)
```

```{r}
glimpse(agpop_mutated)
```
:::

# 3 Plotting Ternary Diagram with R

## 3.1 Plotting static ternary diagrams

### 3.1.1 Basic plot with Size Mappings

**Adding Aesthetics** A basic plot tells us the distribution, but not the scale. By mapping the size of the dots to the `TOTAL` population, we immediately see which subzones represent the majority of the population.

```{r}
ggtern(data = agpop_mutated,
       aes(x = YOUNG,
           y = ACTIVE, 
           z = OLD)) +
  geom_point(aes(size = TOTAL), 
             color = "#0072B2", 
             alpha = 0.6) + 
  scale_size_continuous(range = c(1, 10)) + 
  labs(title = "Population Structure by Subzone, 2018",
       size = "Total Population") +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust=0.5, face="bold"))
```

### 3.1.2 Adding Directional Axes

**Adding Axis Arrows for Readability** Ternary plots can be tricky to read. By adding `theme_showarrows()`, we generate directional arrows along the edges, teaching the viewer how to read the chart.

```{r}
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  geom_point(aes(size = TOTAL), alpha = 0.5, color = "darkred") +
  labs(title = "Population Structure with Directional Axes",
       size = "Total Population") +
  theme_bw() +
  theme_showarrows() +
  theme_clockwise() +
  theme(plot.title = element_text(hjust=0.5, face="bold"),
        legend.position = "bottom")
```

### 3.1.3 Density Contours

**Density Contours to Solve Overplotting** When you have hundreds of subzones, dots overlap. Adding density contours (`stat_density_tern`) acts like a heatmap, showing the highest concentration of data points.

```{r}
ggtern(data = agpop_mutated, aes(x = YOUNG, y = ACTIVE, z = OLD)) +
  stat_density_tern(geom = 'polygon', 
                    aes(fill = ..level.., alpha = ..level..), 
                    show.legend = FALSE) +
  geom_point(size = 1.5, color = "black", alpha = 0.5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Density of Age Structures, 2018") +
  theme_bw() +
  theme(plot.title = element_text(hjust=0.5, face="bold"))
```

### 3.1.4 Plotting 2 groups & Tricolore package

Let's prepare data from 2000 and 2018 so we can compare the changes.

```{r}
agpop_mutated2 <- pop_data %>%
  mutate(`Year` = as.character(Year)) %>%
  pivot_wider(names_from = AG, values_from = Population) %>%
  mutate(YOUNG = rowSums(across(`0_to_4`:`20_to_24`)),
         ACTIVE = rowSums(across(`25_to_29`:`60_to_64`)),
         OLD = rowSums(across(`65_to_69`:`90_and_over`)),
         TOTAL = rowSums(across(YOUNG:OLD))) %>%
  filter(Year %in% c(2018, 2000)) %>%         
  filter(TOTAL > 0)
```

Loading additional functions for [Tricolore](https://github.com/ikashnitsky/compare-pop-eu-us/blob/master/compare-eu-us.R).

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# coordinates and labels for the centered gridlines of a ternary diagram
TernaryCentroidGrid <- function (center) {
    labels <- seq(-1, 1, 0.1)
    labels <- data.frame(
        L = labels[labels >= -center[1]],
        T = labels[labels >= -center[2]],
        R = labels[labels >= -center[3]]
    )
    breaks = data.frame(
        L = labels$L + center[1],
        T = labels$T + center[2],
        R = labels$R + center[3]
    )
    list(labels = labels, breaks = breaks)
}

# a function to create zooming limits
zoom_limits <- function(df, keep_center = TRUE, one_pp_margin = FALSE, center = apply(df, 2, mean, na.rm = T)) {
    mins <- apply(df, 2, min)
    span <- max(apply(df, 2, function(x) diff(range(x))))
    if(one_pp_margin == TRUE & min(mins) > .01){
        mins <- mins - .01
        span <- span + .01
    }
    if(keep_center == TRUE){
        limits <- rbind(center - (1/3)*span/(sqrt(2)/2), center + (2/3)*span/(sqrt(2)/2))
    } else {
        limits <- rbind(mins, c(1 - (mins[2] + mins[3]), 1 - (mins[1] + mins[3]), 1 - (mins[1] + mins[2]))) 
    }
    return(limits)
}
```

Adding ternary colours hexcodes as a column:

```{r}
#| code-fold: true
#| code-summary: "Show the code"

center <- agpop_mutated2 %>% 
    select("YOUNG", "ACTIVE", "OLD") %>% 
    summarise_all(.funs = mean) %>% 
    gather() %>% 
  pull(value)

mins <- apply(agpop_mutated2 %>% select("YOUNG", "ACTIVE", "OLD"), 2, min)

zommed_side <- (1 - (mins[2] + mins[3])) - mins[1]
true_scale <- 1 / zommed_side

tric <- Tricolore(
    agpop_mutated2, p1 = 'YOUNG', p2 = 'ACTIVE', p3 = 'OLD',
    center = NA, show_data = FALSE, spread = true_scale,
    contrast = .5, lightness = 1, chroma = 1, hue = 2/12,
    breaks = 20,
    crop = TRUE, label_as = "pct_diff"
)

agpop_mutated2$hex <- tric$rgb
legend_grid <- TernaryCentroidGrid(center)
legend_limits <- zoom_limits(
    df = agpop_mutated2 %>% select("YOUNG", "ACTIVE", "OLD"),
    keep_center = FALSE,
    one_pp_margin = TRUE
)
```

Plotting for Year 2000 and 2018:

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Plot 2000
tric$key +
    geom_point(data = agpop_mutated2 %>% filter(Year == "2000"), aes(YOUNG, ACTIVE, z = OLD), 
               shape = 21, fill = "grey50", size = .5, alpha = .5)+
    geom_point(data = tibble(YOUNG = center[1], ACTIVE = center[2], OLD = center[3]), 
               aes(YOUNG, ACTIVE, z = OLD), 
               shape = 43, color = "black", size = 5)+
    scale_L_continuous(NULL, limits = legend_limits[,1]) +
    scale_T_continuous(NULL, limits = legend_limits[,2]) +
    scale_R_continuous(NULL, limits = legend_limits[,3]) +
    theme_classic() +
    theme(text = element_text(size = 10, color = "grey20"))

# Plot 2018
tric$key +
    geom_point(data = agpop_mutated2 %>% filter(Year == "2018"), aes(YOUNG, ACTIVE, z = OLD), 
               shape = 21, fill = "white", size = .5, alpha = .5)+
    geom_point(data = tibble(YOUNG = center[1], ACTIVE = center[2], OLD = center[3]), 
               aes(YOUNG, ACTIVE, z = OLD),  
               shape = 43, color = "black", size = 5)+
    scale_L_continuous(NULL, limits = legend_limits[,1]) +
    scale_T_continuous(NULL, limits = legend_limits[,2]) +
    scale_R_continuous(NULL, limits = legend_limits[,3]) +
    theme_classic() +
    theme(text = element_text(size = 10, color = "grey20"))
```

## 3.2 Interactive ternary diagram with `plot_ly`

**Rich Interactive Tooltips in Plotly** By modifying the `text` and `hoverinfo` parameters, we create custom tooltips that reveal exactly which geographic area the dot represents, along with its specific demographic makeup.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "black"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating an ENHANCED plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("#2ca02c"), 
  type = "scatterternary",
  text = ~paste("<b>Area:</b>", PA, "<br>",
                "<b>Subzone:</b>", SZ, "<br>",
                "Young:", YOUNG, "<br>",
                "Active:", ACTIVE, "<br>",
                "Old:", OLD, "<br>",
                "Total:", TOTAL),
  hoverinfo = "text",
  marker = list(size = 7, opacity = 0.6, line = list(color = "black", width = 0.5))
) %>%
  layout(
    annotations = label("Ternary Interactive"), 
    ternary = ternaryAxes
  )
```

## 3.3 Animated Ternary Plot

### 3.3.1 Using gganimate

::: callout-note
The section below attempts to create a animated ternary plot using gganimate(), but it seems like there is some [problem](https://stackoverflow.com/questions/67015151/animate-a-ternary-plot) getting ggtern to work with gganimate.
:::

```{r}
agpop_mutated3 <- pop_data %>%
  pivot_wider(names_from = AG, values_from = Population) %>%
  mutate(YOUNG = rowSums(across(`0_to_4`:`20_to_24`)),
         ACTIVE = rowSums(across(`25_to_29`:`60_to_64`)),
         OLD = rowSums(across(`65_to_69`:`90_and_over`)),
         TOTAL = rowSums(across(YOUNG:OLD))) %>%
  filter(TOTAL > 0)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

td2 <- ggtern(data = agpop_mutated3,
           aes(x = YOUNG, 
               y = ACTIVE, 
               z = OLD,
               size = TOTAL,
               color= PA))+
    geom_point(alpha =0.7, show.legend=FALSE)+
    scale_color_viridis_d(option="plasma")+
    scale_size(range= c(2,15)) +
    scale_L_continuous("Young\n(0-24)", ) +
    scale_T_continuous("Working age\n(25-64)", ) +
    scale_R_continuous("Aged\n(65+)", )+
    labs(title="Comparing Population Age Structures from 2000 to 2018",
         subtitles='Year: {frame_time}')+
    Larrowlab("% aged 0-24") +
    Tarrowlab("% aged 25-64") +
    Rarrowlab("% aged 65+") +
    theme_bw() +
    theme(tern.axis.arrow.show = TRUE, 
          tern.axis.ticks.length.major = unit(9, "pt"),
          tern.axis.text = element_text(size = 8, colour = "grey20"),
          tern.axis.title.T = element_text(),
          tern.axis.title.L = element_text(hjust = 0.2, vjust = 0.7, angle = -60),
          tern.axis.title.R = element_text(hjust = 0.8, vjust = 0.6, angle = 60),
          text = element_text(size = 10, color = "grey20"),
          legend.position = "none",
          plot.title = element_text(hjust=0.5, face="bold"),
          plot.subtitle = element_text(hjust=0.5, face="bold")
)

td2
```

### 3.3.2 Using plotly

```{r}
#| code-fold: true
#| code-summary: "Show the code"
td3 <- plot_ly(agpop_mutated3, 
               a=agpop_mutated3$ACTIVE, 
               b=agpop_mutated3$YOUNG, 
               c=agpop_mutated3$OLD,
                    frame = agpop_mutated3$Year,
                    mode = 'markers',
                    size = agpop_mutated3$TOTAL,
                    color = agpop_mutated3$PA,
                    type = "scatterternary",
                    text = ~paste('Young:',sep='', YOUNG,
                                  '<br>Economically Active:', ACTIVE,
                                  '<br>Aged:',OLD,
                                  '<br>Subzone:', SZ, 
                                  '<br>Planning Area:', PA),
                    hoverinfo="text",
                    marker = list(symbol = 'circle', opacity=0.8,
                                  sizemode="diameter",sizeref=2,
                                  line = list(width = 2, color = '#FFFFFF'))) %>%
                    layout(
                            title = 'Demographic Composition of Singapore',
                            ternary=list(aaxis=list(title="Economically Active"),
                                         baxis = list(title="Young"),
                                         caxis = list(title="Aged")),
                            autosize = FALSE,
                            automargin = FALSE,
                            margin = list(l=50,r=50,b=100,t=100,pad=4),
                            width = 600, 
                            height = 500) 
td3
```
