---
format:
  revealjs:
    theme: simple
    slide-number: true
    show-slide-number: all
    chalkboard: true
    preview-links: auto
    width: 1980
    height: 1080
execute:
  echo: false
  warning: false
  message: false
---

```{r}
pacman::p_load(
  tidyverse, readxl, knitr, scales,
  patchwork, ggdist, ggridges,
  ggstatsplot, viridis, broom
)

if ("package:ggtern" %in% search()) {
  detach("package:ggtern", unload = TRUE, character.only = TRUE)
}

theme_set(ggplot2::theme_minimal(base_size = 15))

theme_update(
  text = element_text(color = "#1F2937"),
  plot.background = element_rect(fill = "white", color = NA),
  panel.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_line(color = "#E5E7EB", linewidth = 0.35),
  panel.grid.minor = element_blank(),
  axis.text = element_text(color = "#4B5563", size = 9),
  axis.title = element_text(color = "#111827", face = "bold", size = 10),
  plot.title = element_text(color = "#111827", face = "bold", size = 16),
  plot.subtitle = element_text(color = "#4B5563", size = 11),
  legend.position = "bottom",
  legend.background = element_rect(fill = "white", color = NA),
  legend.text = element_text(color = "#374151", size = 8),
  legend.title = element_text(color = "#111827", face = "bold", size = 9),
  strip.text = element_text(color = "#111827", face = "bold", size = 10)
)

theme_update(
  tern.axis.ticks.length.major = rel(0.5),
  tern.axis.ticks.length.minor = rel(0.25)
)

safe_divide <- function(numerator, denominator) {
  if_else(denominator > 0, numerator / denominator, NA_real_)
}

premium_brands <- c(
  "AUDI", "BMW", "FERRARI", "HUMMER", "INFINITI", "JAGUAR",
  "LAND ROVER", "LEXUS", "MASERATI", "MERCEDES", "MINI",
  "PORSCHE", "SAAB", "VOLVO"
)

portfolio <- read_delim(
  file = "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  show_col_types = FALSE
) |>
  select(
    insured_id, year,
    policy_type, policy_status, business_type, payment_frequency, bonus_score,
    driver_age, vehicle_age, fuel_type, vehicle_value, power_to_weight_ratio,
    vehicle_brand, municipality_type, circulation_area,
    total_premium, total_claims, total_incurred, total_exposure
  ) |>
  mutate(
    policy_type = recode(
      policy_type,
      "TP" = "Third party",
      "TPG" = "Third party + glass",
      "CC" = "Third party + add-ons",
      "COMP_E" = "Comprehensive with excess",
      "COMP_N" = "Comprehensive without excess"
    ),
    policy_type = factor(
      policy_type,
      levels = c(
        "Third party", "Third party + glass", "Third party + add-ons",
        "Comprehensive with excess", "Comprehensive without excess"
      )
    ),
    policy_status = recode(policy_status, "A" = "Active", "C" = "Cancelled"),
    business_type = recode(business_type, "NB" = "New business", "P" = "Portfolio"),
    bonus_score = recode(bonus_score, "G" = "Good", "N" = "Neutral", "B" = "Bad"),
    bonus_score = factor(bonus_score, levels = c("Good", "Neutral", "Bad")),
    fuel_type = recode(fuel_type, "D" = "Diesel", "G" = "Gasoline", .missing = "Unknown"),
    municipality_type = recode(municipality_type, "I" = "Inland", "C" = "Coastal", "IS" = "Islands"),
    circulation_area = recode(circulation_area, "U" = "Urban", "R" = "Rural"),
    vehicle_brand = replace_na(vehicle_brand, "Unknown"),
    brand_tier = if_else(vehicle_brand %in% premium_brands, "Premium brand", "Mainstream brand"),
    geo_segment = str_c(municipality_type, " - ", circulation_area),
    loss_ratio = safe_divide(total_incurred, total_premium),
    underwriting_result = total_premium - total_incurred,
    claim_flag = total_claims > 0,
    claim_frequency = safe_divide(total_claims, total_exposure),
    claim_severity = safe_divide(total_incurred, total_claims),
    profit_status = case_when(
      total_claims == 0 ~ "No claim",
      loss_ratio <= 1 ~ "Claim, still profitable",
      TRUE ~ "Loss-making"
    ),
    profit_status = factor(
      profit_status,
      levels = c("Loss-making", "Claim, still profitable", "No claim")
    ),
    exposure_band = cut(
      total_exposure,
      breaks = c(0, 0.35, 0.75, 0.95, Inf),
      labels = c("Very short", "Short", "Medium", "Full"),
      include.lowest = TRUE
    ),
    driver_age_band = cut(
      driver_age,
      breaks = c(17, 25, 35, 45, 55, 65, Inf),
      labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "66+")
    ),
    vehicle_age_band = cut(
      vehicle_age,
      breaks = c(-Inf, 3, 7, 12, 20, Inf),
      labels = c("0-3", "4-7", "8-12", "13-20", "21+")
    ),
    vehicle_value_band = ntile(vehicle_value, 5),
    vehicle_value_band = factor(vehicle_value_band, labels = c("Lowest", "Low", "Middle", "High", "Highest")),
    same_insured_years = ave(year, insured_id, FUN = function(x) length(unique(x))),
    loyalty_band = case_when(
      same_insured_years >= 3 ~ "Observed all 3 years",
      same_insured_years == 2 ~ "Observed 2 years",
      TRUE ~ "Observed 1 year"
    ),
    loyalty_band = factor(loyalty_band, levels = c("Observed 1 year", "Observed 2 years", "Observed all 3 years")),
    across(c(policy_status, business_type, fuel_type, municipality_type, circulation_area, brand_tier, geo_segment), as_factor)
  )

loss_ratio_cap_99 <- quantile(
  portfolio$loss_ratio[is.finite(portfolio$loss_ratio)],
  probs = 0.99,
  na.rm = TRUE
)

dictionary_raw <- read_excel("data/Descriptive of variables.xlsx")

portfolio <- portfolio |>
  mutate(loss_ratio_capped_99 = pmin(loss_ratio, loss_ratio_cap_99))

segment_summary <- function(data, segment_var) {
  data |>
    filter(!is.na({{ segment_var }})) |>
    group_by({{ segment_var }}) |>
    summarise(
      policies = n(),
      exposure = sum(total_exposure, na.rm = TRUE),
      premium = sum(total_premium, na.rm = TRUE),
      claims = sum(total_claims, na.rm = TRUE),
      incurred = sum(total_incurred, na.rm = TRUE),
      underwriting_result = sum(underwriting_result, na.rm = TRUE),
      loss_ratio = incurred / premium,
      claim_frequency = claims / exposure,
      claim_severity = incurred / claims,
      .groups = "drop"
    ) |>
    arrange(desc(premium))
}

portfolio_summary <- portfolio |>
  summarise(
    policy_years = n(),
    insured_policies = n_distinct(insured_id),
    exposure = sum(total_exposure, na.rm = TRUE),
    premium = sum(total_premium, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    incurred = sum(total_incurred, na.rm = TRUE),
    underwriting_result = sum(underwriting_result, na.rm = TRUE),
    loss_ratio = incurred / premium,
    claim_frequency = claims / exposure,
    claim_severity = incurred / claims
  )

year_summary <- segment_summary(portfolio, year)
policy_type_summary <- segment_summary(portfolio, policy_type)
business_summary <- segment_summary(portfolio, business_type)
bonus_summary <- segment_summary(portfolio, bonus_score)
exposure_summary <- segment_summary(portfolio, exposure_band)
driver_age_summary <- segment_summary(portfolio, driver_age_band)
vehicle_age_summary <- segment_summary(portfolio, vehicle_age_band)
vehicle_value_summary <- segment_summary(portfolio, vehicle_value_band)
brand_tier_summary <- segment_summary(portfolio, brand_tier)
loyalty_summary <- segment_summary(portfolio, loyalty_band)
geo_summary <- segment_summary(portfolio, geo_segment)
```

## Take-home Exercise 1: Visual Analytics for Data Discovery

### Mythbusting Risk Stereotypes in Motor Insurance

**Method:** 10 mythbusting research questions, tidyverse data wrangling, and best-fit ggplot visual analytics

::: notes
I am framing this presentation as a mythbusting investigation of motor insurance risk. Instead of assuming common stereotypes are true, I test them against portfolio data from 2022 to 2024. I focus on loss ratio, claim frequency, claim severity, underwriting result, exposure, and retention because these measures show whether a segment is actually profitable, risky, or worth protecting.
:::

## Executive Summary

The portfolio is profitable overall, but profitability is uneven.

|  |  |  |
|------------------------|------------------------|------------------------|
| **Strategic objective** | **Main finding** | **Business action** |
| Profitability | Comprehensive without excess and very short-exposure policies create the clearest pressure. | Review excess design, short-duration loading, and claims controls. |
| Retention | Customers observed all 3 years are materially more profitable than one-year customers. | Protect low-risk loyal customers from blunt repricing. |
| Calibrated pricing | Some stereotypes are real, but many are mixed. | Price using frequency, severity, materiality, and evidence together. |

::: notes
My overall message is that the portfolio is profitable, but the profitability is uneven. I want to show that stereotypes can be useful starting hypotheses, but they should not become pricing rules by themselves. The clearest actions are to review comprehensive without excess, monitor very short-exposure policies, retain profitable loyal customers, and use age, brand, vehicle value, and geography as calibrated supporting signals.
:::

## Contents

1.  Executive summary and objectives

2.  Methodology, data, and technique rationale

3.  Portfolio trend and policy-type triage

4.  Mythbusting findings by risk dimension

5.  Answers to the 10 research questions

6.  Recommendations

::: notes
I will first explain the objective, data preparation, and technique choices, then move into the key visual findings. The flow is designed to move from overall portfolio performance into segment-level mythbusting. I end by answering the 10 research questions and linking the evidence back to profitability, retention, and calibrated pricing.
:::

## Methodology and Data

**Core question:** which insurance stereotypes are real, and which are myths?

The strategic objectives are to:

-   maximise overall portfolio profitability;
-   maximise retention and loyalty among low-risk, highly profitable customers;
-   identify high-risk segments for appropriately calibrated premium pricing.

The analysis answers 10 research questions covering:

-   conventional and demographic factors;
-   brand-and-status factors;
-   behavioural and claim-history factors;
-   loyalty and policy-status factors;
-   geographic frequency and severity differences.

## Technique Rationale

|  |  |
|------------------------------------|------------------------------------|
| **Decision need** | **Visual technique** |
| Profitability pressure | KPI tiles with embedded `geom_text()` labels |
| Frequency vs severity diagnosis | KPI matrices and `stat_pointinterval()` |
| Retention and loyalty value | KPI bars by observed customer duration |
| Stereotype overlap | Half-eye distributions and boxplots |
| Hidden short-exposure risk | Ridgelines with quantile shading |
| Interaction risk | Heatmaps with embedded labels |

```{r}
portfolio_summary |>
  transmute(
    `Policy-years` = comma(policy_years),
    `Insured policies` = comma(insured_policies),
    Exposure = comma(exposure, accuracy = 1),
    Premium = dollar(premium, prefix = ""),
    Incurred = dollar(incurred, prefix = ""),
    `Loss ratio` = percent(loss_ratio, accuracy = 0.1),
    `Claim frequency` = number(claim_frequency, accuracy = 0.001)
  ) |>
  kable()
```

::: notes
Here I am showing how the analysis is made reproducible. The sandbox starts from the raw CSV and Excel dictionary, then derives loss ratio, frequency, severity, exposure bands, age bands, loyalty, brand tier, and geography. I also match each visual technique to a decision need: tiles summarise KPI pressure, half-eye plots show distribution overlap, ridgelines show tail risk, and heatmaps reveal interaction effects.
:::

## One-Look Portfolio Trend

```{r}
#| fig-width: 12
#| fig-height: 5.5

year_long <- year_summary |>
  select(year, premium, incurred, underwriting_result, loss_ratio, claim_frequency) |>
  pivot_longer(
    cols = c(premium, incurred, underwriting_result, loss_ratio, claim_frequency),
    names_to = "measure",
    values_to = "value"
  ) |>
  mutate(
    measure = factor(
      measure,
      levels = c("claim_frequency", "loss_ratio", "underwriting_result", "incurred", "premium"),
      labels = c("Claim Frequency", "Loss Ratio", "Underwriting Result", "Incurred Loss", "Premium")
    ),
    label = case_when(
      measure == "Loss Ratio" ~ percent(value, accuracy = 0.1),
      measure == "Claim Frequency" ~ number(value, accuracy = 0.001),
      TRUE ~ label_number(scale_cut = cut_short_scale())(value)
    )
  )

ggplot(year_long, aes(x = factor(year), y = "1", fill = value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(
    aes(
      label = label, 
      colour = value > median(value) 
    ), 
    size = 3.2, 
    fontface = "bold"
  ) +
  scale_colour_manual(values = c("TRUE" = "#111827", "FALSE" = "#FFFFFF"), guide = "none") +
  facet_grid(measure ~ ., switch = "y") +
  scale_fill_viridis_c(option = "C", guide = "none") +
  theme(
    axis.text.y = element_blank(),    
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),   
    strip.text.y.left = element_text(angle = 0, hjust = 1, face = "bold", size = 9), 
    strip.placement = "outside",      
    panel.spacing = unit(0.1, "lines") 
  ) +
  labs(
    title = "Portfolio Trend Dashboard", 
    x = "Year", 
    y = NULL
  )
```

::: notes
This slide gives the portfolio-level baseline before I discuss segments. I use a static dashboard so premium, incurred loss, underwriting result, loss ratio, and claim frequency can be compared at once. The main pattern I want the audience to see is whether deterioration is driven by profitability, claim activity, or portfolio scale, rather than treating the loss ratio as a standalone number.
:::

## Policy-Type Triage

```{r}
#| fig-width: 11
#| fig-height: 5.6

triage_matrix_data <- policy_type_summary |>
  mutate(premium_share = premium / sum(premium)) |>
  select(policy_type, claim_frequency, claim_severity, loss_ratio, premium_share) |>
  pivot_longer(-policy_type, names_to = "metric", values_to = "value") |>
  mutate(
    metric = factor(
      metric,
      levels = c("claim_frequency", "claim_severity", "loss_ratio", "premium_share"),
      labels = c("Claim Frequency", "Claim Severity", "Loss Ratio", "Premium Share")
    ),
    label = case_when(
      metric == "Claim Severity" ~ dollar(value, prefix = ""),
      metric == "Claim Frequency" ~ number(value, accuracy = 0.001),
      TRUE ~ percent(value, accuracy = 0.1)
    )
  ) |>
  group_by(metric) |>
  mutate(scaled_value = rescale(value)) |>
  ungroup()

ggplot(triage_matrix_data, aes(x = metric, y = policy_type, fill = scaled_value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(
    aes(label = label, colour = scaled_value > 0.6),
    size = 3,
    fontface = "bold"
  ) +
  scale_colour_manual(values = c("TRUE" = "#111827", "FALSE" = "#FFFFFF"), guide = "none") +
  scale_fill_viridis_c(option = "C", guide = "none") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold", size = 8),
    axis.text.y = element_text(size = 8),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  labs(
    title = "Policy-Type Triage Matrix",
    subtitle = "Color intensity indicates relative highs and lows within each specific metric",
    x = NULL,
    y = NULL
  )
```

::: notes
This slide is my first prioritisation view. I compare policy type across loss ratio, claim frequency, claim severity, and premium share. The important point is that not every weak segment needs the same response. A high-frequency segment suggests product, deductible, or renewal controls, while a high-severity segment suggests reserve, claims, or coverage review.
:::

## Profitability Status

```{r}
#| fig-width: 10
#| fig-height: 5.6

ggbarstats(
  data = portfolio,
  x = policy_type,
  y = profit_status,
  label = "percentage",
  results.subtitle = FALSE,
  ggtheme = ggplot2::theme_get(),
  title = "Policy-year profitability status by policy type",
  xlab = "Policy type",
  ylab = "Policy-year records",
  legend.title = "Profitability status"
) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 8))
```

::: notes
Here I break each policy type into no-claim, claim-but-profitable, and loss-making records. This helps me show whether a product problem is broad-based or concentrated in a loss-making tail. The key interpretation is that comprehensive without excess should be treated as a frequency-management concern, and we should not assume that higher cover automatically means stronger profitability.
:::

## Bonus Score Signal

```{r}
#| fig-width: 10
#| fig-height: 5.6

ggplot(
  portfolio |> filter(!is.na(bonus_score), loss_ratio_capped_99 > 0),
  aes(x = bonus_score, y = loss_ratio_capped_99, fill = bonus_score)
) +
  stat_halfeye(
    adjust = 0.8, 
    width = 0.55, 
    justification = -0.2, 
    .width = 0, 
    alpha = 0.65, 
    show.legend = FALSE
  ) +
  geom_boxplot(
    width = 0.16, 
    outlier.shape = NA, 
    alpha = 0.5, 
    show.legend = FALSE
  ) +
  geom_jitter(
    width = 0.08, 
    alpha = 0.05, 
    size = 0.3, 
    show.legend = FALSE,
    color = "#64748B" 
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#EF4444", linewidth = 1) +
  scale_y_continuous(labels = number_format(accuracy = 0.1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Bonus Score Myth Check: Claim Severity",
    subtitle = "Distribution of loss ratios (excluding zero-claim policies); dashed line = break-even",
    x = "Bonus Score",
    y = "Loss Ratio (Capped at 99th percentile)"
  )
```

::: notes
This slide tests whether bonus score cleanly separates good and bad risks. The half-eye plot shows the shape and overlap of the loss-ratio distributions, while the boxplot summarises the central spread. My interpretation is that bonus score is directionally useful, especially for frequency, but the overlap means it should support pricing rather than act as a single blunt rule.
:::

## Exposure Duration and Loyalty

```{r}
#| fig-width: 12
#| fig-height: 5.8

p_exposure <- ggplot(
  portfolio |> filter(claim_flag, !is.na(exposure_band), !is.na(loss_ratio_capped_99)),
  aes(x = loss_ratio_capped_99, y = exposure_band, fill = factor(after_stat(quantile)))
) +
  geom_density_ridges_gradient(
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE,
    scale = 1.6,
    rel_min_height = 0.005,
    colour = "white",   
    linewidth = 0.35
  ) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#EF4444", linewidth = 1) +
  scale_fill_viridis_d(option = "C", guide = "none") +
  theme(plot.margin = margin(t = 10, r = 20, b = 10, l = 10)) +
  labs(title = "Claimant LR by duration", x = "Loss ratio", y = NULL)

p_loyalty <- loyalty_summary |>
  mutate(premium_share = premium / sum(premium)) |>
  select(loyalty_band, premium_share, loss_ratio) |>
  pivot_longer(-loyalty_band, names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, "premium_share" = "Premium share", "loss_ratio" = "Loss ratio")) |>
  ggplot(aes(x = metric, y = loyalty_band, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(
      label = percent(value, accuracy = 0.1),
      colour = value > 0.70 
    ),
    size = 3.2,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_colour_manual(values = c("TRUE" = "#111827", "FALSE" = "#FFFFFF")) +
  scale_fill_viridis_c(option = "C", guide = "none") +
  theme(
    panel.grid = element_blank(), 
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(title = "Retention value", x = NULL, y = NULL)

final_plot <- p_exposure + p_loyalty + plot_layout(widths = c(1.3, 1))

final_plot
```

::: notes
This slide connects two strategic questions: early-duration risk and customer loyalty. On the left, I use ridgelines to check whether short-exposure policies have a heavier adverse loss-ratio distribution. On the right, I compare premium share and loss ratio by observed customer duration. The pattern supports monitoring short-exposure policies while protecting longer-observed customers that show stronger profitability.
:::

## Cancellation Quality

```{r}
#| fig-width: 11
#| fig-height: 5.6

ggbarstats(
  data = portfolio |> filter(!is.na(policy_status), !is.na(bonus_score)),
  x = bonus_score,
  y = policy_status,
  label = "percentage",
  results.subtitle = FALSE,
  ggtheme = ggplot2::theme_get(),
  title = "Cancellation quality by bonus score",
  subtitle = "Cancellation is useful only if poor risks leave faster than good risks",
  xlab = "Bonus score",
  ylab = "Policy-year records",
  legend.title = "Policy status"
)
```

::: notes
Here I explain why cancellation should not be interpreted too casually. If high-risk policies cancel, attrition may reduce pressure, but if profitable customers leave, retention quality worsens. I use the bonus-score split to show whether weaker risks are actually leaving. My main takeaway is that the insurer should detect poor risks earlier instead of relying on cancellation to clean the portfolio.
:::

## Driver and Vehicle Interaction

```{r}
#| fig-width: 10
#| fig-height: 5.8

age_heatmap <- portfolio |>
  filter(!is.na(driver_age_band), !is.na(vehicle_age_band)) |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(
    exposure = sum(total_exposure, na.rm = TRUE),
    claims = sum(total_claims, na.rm = TRUE),
    claim_frequency = if_else(exposure > 0, claims / exposure, 0),
    .groups = "drop"
  )

ggplot(age_heatmap, aes(x = vehicle_age_band, y = driver_age_band, fill = claim_frequency)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(
      label = number(claim_frequency, accuracy = 0.01),
      colour = claim_frequency > 0.50
    ), 
    size = 3.2, 
    fontface = "bold"
  ) +
  scale_colour_manual(values = c("TRUE" = "#111827", "FALSE" = "#FFFFFF"), guide = "none") +
  scale_fill_viridis_c(
    option = "C", 
    limits = c(0, 0.75), 
    oob = scales::squish, 
    guide = "none"
  ) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  labs(
    title = "Claim frequency interaction: Driver Age x Vehicle Age",
    subtitle = "Color scale capped at 0.75 to prevent low-exposure outliers from washing out the gradient",
    x = "Vehicle Age Band",
    y = "Driver Age Band"
  )
```

::: notes
This heatmap shows why I include multivariate analysis. Looking only at driver age or vehicle age separately can hide important combinations. Each cell shows claim frequency for a driver-age and vehicle-age pairing. The cells with higher frequency indicate where pricing interactions or underwriting rules may be more useful than a simple age-only or vehicle-age-only adjustment.
:::

## Brand, Value, and Geography

```{r}
#| fig-width: 12
#| fig-height: 5.8

p_brand <- ggplot(
  portfolio |> filter(claim_flag, !is.na(brand_tier), !is.na(claim_severity), claim_severity > 0),
  aes(x = brand_tier, y = claim_severity)
) +
  stat_pointinterval(.width = c(0.8, 0.95), point_size = 2.6, aes(colour = factor(after_stat(.width)))) +
  scale_y_log10(labels = dollar_format(prefix = "")) +
  scale_colour_manual(values = c("0.8" = "#4C78A8", "0.95" = "#F58518"), guide = "none") +
  labs(title = "Severity by brand", x = NULL, y = "Claim severity")

p_value <- ggplot(
  vehicle_value_summary,
  aes(x = vehicle_value_band, y = "Loss ratio", fill = loss_ratio)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = percent(loss_ratio, accuracy = 0.1)), size = 2.8, colour = "white", fontface = "bold") +
  scale_fill_viridis_c(option = "C", guide = "none") +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
    axis.text.y = element_blank(),                   
    panel.grid = element_blank()                     
  ) +
  labs(title = "LR by vehicle value", x = NULL, y = NULL)

p_geo <- ggplot(
  geo_summary,
  aes(x = fct_reorder(geo_segment, loss_ratio), y = "Loss ratio", fill = loss_ratio)
) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = percent(loss_ratio, accuracy = 0.1)), size = 2.7, colour = "white", fontface = "bold") +
  scale_fill_viridis_c(option = "C", guide = "none") + 
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 6.5),
    axis.text.y = element_blank(),                 
    panel.grid = element_blank()                       
  ) +
  labs(title = "LR by geography", x = NULL, y = NULL)

final_plot <- p_brand | p_value | p_geo

final_plot
```

::: notes
This slide tests three common status and location stereotypes. I use point intervals for brand severity because uncertainty matters, and tile views for vehicle value and geography so the loss-ratio differences are easy to compare. My interpretation is that these variables are useful supporting lenses, but they should lead to calibrated loadings, not automatic surcharges.
:::

## Answers to the 10 Research Questions

```{r}
driver_young <- driver_age_summary |> filter(driver_age_band == "18-25")
driver_loss_top <- driver_age_summary |> slice_max(loss_ratio, n = 1, with_ties = FALSE)
vehicle_new <- vehicle_age_summary |> filter(vehicle_age_band == "0-3")
vehicle_loss_top <- vehicle_age_summary |> slice_max(loss_ratio, n = 1, with_ties = FALSE)
comp_no_excess <- policy_type_summary |> filter(policy_type == "Comprehensive without excess")
premium_brand <- brand_tier_summary |> filter(brand_tier == "Premium brand")
mainstream_brand <- brand_tier_summary |> filter(brand_tier == "Mainstream brand")
value_highest <- vehicle_value_summary |> filter(vehicle_value_band == "Highest")
value_lowest <- vehicle_value_summary |> filter(vehicle_value_band == "Lowest")
bonus_bad <- bonus_summary |> filter(bonus_score == "Bad")
bonus_neutral <- bonus_summary |> filter(bonus_score == "Neutral")
exposure_very_short <- exposure_summary |> filter(exposure_band == "Very short")
loyal_3yr <- loyalty_summary |> filter(loyalty_band == "Observed all 3 years")
loyal_1yr <- loyalty_summary |> filter(loyalty_band == "Observed 1 year")
status_summary <- segment_summary(portfolio, policy_status)
status_cancelled <- status_summary |> filter(policy_status == "Cancelled")
status_active <- status_summary |> filter(policy_status == "Active")
area_summary <- segment_summary(portfolio, circulation_area)
area_urban <- area_summary |> filter(circulation_area == "Urban")
area_rural <- area_summary |> filter(circulation_area == "Rural")

tibble(
  RQ = paste0("RQ", 1:10),
  Verdict = c("Mixed", "Partly real", "Busted", "Mostly busted", "Mixed", "Mostly real", "Busted", "Busted", "Real", "Mixed"),
  Answer = c(
    str_glue("18-25 has highest frequency ({number(driver_young$claim_frequency, accuracy = 0.001)}) but only {percent(driver_young$loss_ratio, accuracy = 0.1)} LR; worst age LR is {driver_loss_top$driver_age_band}."),
    str_glue("0-3 vehicles claim most often ({number(vehicle_new$claim_frequency, accuracy = 0.001)}), but worst LR is {vehicle_loss_top$vehicle_age_band} ({percent(vehicle_loss_top$loss_ratio, accuracy = 0.1)})."),
    str_glue("Comprehensive without excess has {percent(comp_no_excess$loss_ratio, accuracy = 0.1)} LR and {number(comp_no_excess$claim_frequency, accuracy = 0.001)} frequency."),
    str_glue("Premium brands have higher frequency but lower LR ({percent(premium_brand$loss_ratio, accuracy = 0.1)}) than mainstream brands ({percent(mainstream_brand$loss_ratio, accuracy = 0.1)})."),
    str_glue("Highest-value LR is {percent(value_highest$loss_ratio, accuracy = 0.1)}, but lowest-value LR is also weak at {percent(value_lowest$loss_ratio, accuracy = 0.1)}."),
    str_glue("Bad has highest frequency ({number(bonus_bad$claim_frequency, accuracy = 0.001)}); Neutral has highest LR ({percent(bonus_neutral$loss_ratio, accuracy = 0.1)})."),
    str_glue("Very short exposure is above break-even at {percent(exposure_very_short$loss_ratio, accuracy = 0.1)} LR."),
    str_glue("Observed all 3 years has best LR ({percent(loyal_3yr$loss_ratio, accuracy = 0.1)}) versus one-year customers ({percent(loyal_1yr$loss_ratio, accuracy = 0.1)})."),
    str_glue("Cancelled policies are worse: {percent(status_cancelled$loss_ratio, accuracy = 0.1)} LR versus active {percent(status_active$loss_ratio, accuracy = 0.1)}."),
    str_glue("Urban has higher frequency and LR; rural has higher severity ({dollar(area_rural$claim_severity, prefix = '')}).")
  )
) |>
  kable()
```

::: notes
This slide summarises the 10 research questions as evidence-backed verdicts. I use "mixed", "busted", or "real" to make the conclusions easy to follow. The important point is that a stereotype may be true for frequency but false for overall profitability, so I interpret each myth using loss ratio, frequency, severity, materiality, and loyalty evidence together.
:::

## Recommendations
1.  Retain longer-observed customers with strong profitability.

2.  Review comprehensive without excess and very short-exposure policies first.

3.  Detect cancellation-prone poor risks earlier, before losses emerge.

4.  Use age, brand, vehicle value, and geography as calibrated signals, not stereotypes.

5.  Treat the priority queue as a review list, not as a pricing model.

::: notes
I close by translating the visual findings into actions. For profitability, I would review comprehensive without excess and very short-exposure policies first. For retention, I would protect longer-observed profitable customers. For pricing, I would use the mythbusting evidence to calibrate age, brand, value, geography, and cancellation signals instead of applying broad stereotype-based rules.
:::
