---
title: "Hands-on Exercise 8:"
subtitle: "Visualising and Analysing Time-Oriented Data"
date: "11 June 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1 Overview

This hands-on exercise follows [Chapter 17: Visualising and Analysing Time-oriented Data](https://r4va.netlify.app/chap17) from *R for Visual Analytics*. Rather than reproducing the reference work line-for-line, I rebuild each technique with my **own variations** (different focal series, colour systems, faceting and themes) and then add an **Extras** section with five pieces of analysis that appear in neither the chapter nor the reference notebook.

The techniques covered are:

-   plotting a **calendar heatmap** with `ggplot2` (my version uses a `viridis` colour scale and a Monday-first weekday order),
-   plotting a **cycle plot** (my version profiles Japan over the full 2000-2019 window),
-   plotting a **slopegraph** with `CGPfunctions` (my version contrasts 1980 vs 2010 and highlights the biggest movers),
-   plus five **extras**: a horizon chart, a 24-hour radial attack clock, a GitHub-style calendar grid, a `gridExtra` side-by-side comparison, and a rice-yield growth ranking.

> **Scope rule.** I only use packages that already appear in the source chapter or the reference notebook. The union of those two package lists is the full toolbox below - nothing else is introduced. Note that `viridis`, `gridExtra` and `data.table` are loaded by the reference but never actually used there; my variations and extras put them to work. `ggHoriPlot` is listed by the source chapter but not used by either file, so the horizon chart in Section 6 is built strictly within the allowed toolbox.

# 2 Getting Started

## 2.1 Loading the required packages

```{r}
# data.table is loaded BEFORE lubridate + tidyverse on purpose: data.table also
# exports wday()/month()/hour()/year(), so loading it first lets the lubridate
# and dplyr versions win the masked names (e.g. wday(week_start=), month(label=)).
pacman::p_load(scales, viridis, ggthemes, gridExtra, readxl, knitr,
               data.table, lubridate, tidyverse, CGPfunctions, ggHoriPlot)
```

```{r}
# autosave targets: plots -> images/, cached frames -> data/rds/
if (!dir.exists("images"))   dir.create("images")
if (!dir.exists("data/rds")) dir.create("data/rds", recursive = TRUE)
```

# 3 Plotting Calendar Heatmap

## 3.1 Importing the data

We import the *eventlog.csv* cyber-attack log (≈200k records, spanning Mar-Dec 2015).

```{r}
attacks <- read_csv("data/eventlog.csv")
kable(head(attacks))
```

## 3.2 Data preparation

Like the reference, I convert each UTC timestamp to its **local** time before pulling out the weekday and hour, because the source country's tile only makes sense in local time. My variation differs in two ways:

-   I replace the superseded `do()` + helper-function pattern with a modern `group_modify()` over each timezone, and
-   I derive the weekday with `lubridate::wday(week_start = 1)` mapped to fixed English labels (locale-independent), ordering the axis **Monday at the top** instead of the reference's Saturday-first order.

```{r}
wkday_levels <- c("Monday", "Tuesday", "Wednesday",
                  "Thursday", "Friday", "Saturday", "Sunday")

attacks_cleaned <- attacks %>%
  group_by(tz) %>%
  group_modify(~ {
    local_times <- ymd_hms(.x$timestamp) %>% with_tz(tzone = .y$tz)
    tibble(
      source_country = .x$source_country,
      wkday = wkday_levels[wday(local_times, week_start = 1)],
      hour  = hour(local_times)
    )
  }) %>%
  ungroup() %>%
  mutate(
    # rev() so Monday is drawn at the TOP of the heatmap
    wkday = factor(wkday, levels = rev(wkday_levels)),
    hour  = factor(hour, levels = 0:23)
  )

# cache the expensive timezone-converted frame
write_rds(attacks_cleaned, "data/rds/attacks_cleaned.rds")
kable(head(attacks_cleaned))
```

## 3.3 Building the calendar heatmap

I aggregate to weekday x hour and draw the tiles. The key visual variation from the reference: instead of a two-point `scale_fill_gradient()` (sky-blue to dark-blue), I use a perceptually-uniform **`viridis` "inferno"** ramp so the busiest cells pop.

```{r}
grouped <- attacks_cleaned %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

p_heat <- ggplot(grouped, aes(hour, wkday, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_viridis(name = "# of attacks", option = "inferno", direction = -1) +
  labs(x = NULL, y = NULL,
       title = "Attacks by weekday and time of day",
       subtitle = "Local time of the source country | viridis (inferno) scale") +
  theme(axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 8),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))

ggsave("images/calendar_heatmap.png", p_heat, width = 8, height = 4, dpi = 150)
p_heat
```

## 3.4 Generating data per country

To compare patterns across the most-targeted countries I rank them by attack volume. My variation widens the focus from the reference's **top 4** to the **top 6** countries.

```{r}
attacks_by_country <- count(attacks_cleaned, source_country) %>%
  mutate(percent = percent(n / sum(n))) %>%
  arrange(desc(n))

kable(head(attacks_by_country, 6))
```

```{r}
top6 <- attacks_by_country$source_country[1:6]

top6_attacks <- attacks_cleaned %>%
  filter(source_country %in% top6) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(source_country, levels = top6)) %>%
  na.omit()
```

## 3.5 Plotting the faceted heatmaps

```{r}
p_heat_facet <- ggplot(top6_attacks, aes(hour, wkday, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_viridis(name = "# of attacks", option = "inferno", direction = -1) +
  facet_wrap(~ source_country, ncol = 3) +
  labs(x = NULL, y = NULL,
       title = "Attacks on the top 6 source countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 6),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))

ggsave("images/calendar_heatmap_top6.png", p_heat_facet, width = 10, height = 5, dpi = 150)
p_heat_facet
```

# 4 Plotting Cycle Plot

## 4.1 Data import

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

## 4.2 Deriving month and year

```{r}
air$month <- factor(month(air$`Month-Year`),
                    levels = 1:12,
                    labels = month.abb,
                    ordered = TRUE)
air$year <- year(ymd(air$`Month-Year`))
```

## 4.3 Extracting the target country

The reference profiles Vietnam from 2010. My variation profiles **Japan** across the **full available window (2000-2019)**, giving 20 points per monthly facet instead of 10.

```{r}
japan <- air %>%
  select(`Japan`, month, year) %>%
  filter(year >= 2000)
```

## 4.4 Computing monthly averages

The dashed reference line is each month's long-run mean.

```{r}
hline.data <- japan %>%
  group_by(month) %>%
  summarise(avgvalue = mean(`Japan`))
```

## 4.5 Plotting the cycle plot

Beyond the country/period swap, my styling differs from the reference: a steel-blue trend line, a firebrick **solid** mean line, `theme_few()` instead of `theme_tufte()`, and comma-formatted counts via `scales::label_comma()`.

```{r}
p_cycle <- ggplot() +
  geom_line(data = japan,
            aes(x = year, y = `Japan`, group = month),
            colour = "steelblue4", linewidth = 0.4) +
  geom_hline(data = hline.data,
             aes(yintercept = avgvalue),
             linetype = "solid", colour = "firebrick", linewidth = 0.4) +
  facet_grid(~ month) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Visitor arrivals from Japan by air, Jan 2000-Dec 2019",
       subtitle = "Black-blue line = yearly values | red line = month's long-run mean",
       x = NULL, y = "No. of visitors") +
  theme_few(base_family = "Helvetica") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.subtitle = element_text(size = 8),
        strip.text = element_text(size = 8))

ggsave("images/cycle_plot_japan.png", p_cycle, width = 10, height = 4, dpi = 150)
p_cycle
```

# 5 Plotting Slope Graph

## 5.1 Data import

```{r}
rice <- read_csv("data/rice.csv")
```

## 5.2 Generating the slopegraph

The reference compares **1961 vs 1980**. My variation compares the most recent 30-year span in the data, **1980 vs 2010**, and instead of colouring every country I grey out the field and **highlight the two steepest climbers** (Vietnam and Myanmar) with `newggslopegraph()`'s named `LineColor` argument.

```{r}
rice_years <- rice %>%
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1980, 2010))

# grey baseline, then accent the two biggest movers
groups <- unique(rice_years$Country)
line_pal <- setNames(rep("gray75", length(groups)), groups)
line_pal["VietNam"] <- "#D7263D"
line_pal["Myanmar"] <- "#1B998B"

p_slope <- rice_years %>%
  newggslopegraph(Year, Yield, Country,
                  Title    = "Rice Yield of Top 11 Asian Countries (1980 vs 2010)",
                  SubTitle = "Vietnam and Myanmar posted the steepest gains over the period",
                  Caption  = "HO_EX08 | CR Tech",
                  LineColor = line_pal,
                  LineThickness = 1,
                  DataTextSize = 3,
                  YTextSize = 3)

ggsave("images/slopegraph_rice.png", p_slope, width = 7, height = 7, dpi = 150)
p_slope
```

# 6 Extras

## Horizon chart of air arrivals (`ggHoriPlot`)

`ggHoriPlot` is listed by the source chapter's package set but neither file builds a horizon chart. A horizon plot folds a series into stacked, colour-banded layers so many time series can be compared in a compact strip. Here I profile monthly arrivals for six markets across 2000-2019.

```{r}
arrivals_long <- air %>%
  select(`Month-Year`, China, Japan, `South Korea`, USA, Australia, India) %>%
  pivot_longer(-`Month-Year`, names_to = "country", values_to = "arrivals") %>%
  mutate(date = as.Date(`Month-Year`))

p_horizon <- ggplot(arrivals_long) +
  geom_horizon(aes(date, arrivals, fill = after_stat(Cutpoints)),
               origin = "midpoint", horizonscale = 6) +
  scale_fill_hcl(palette = "RdBu", reverse = TRUE) +
  facet_grid(country ~ .) +
  theme_few() +
  scale_x_date(expand = c(0, 0), date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Monthly visitor arrivals by air, 2000-2019",
       subtitle = "Horizon bands: blue = below the series midpoint, red = above",
       x = NULL) +
  theme(panel.spacing.y = unit(0, "lines"),
        strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
        axis.text.y  = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        legend.position = "none",
        plot.subtitle = element_text(size = 8))

ggsave("images/extra_horizon.png", p_horizon, width = 9, height = 5, dpi = 150)
p_horizon
```

## A 24-hour radial "attack clock" (`coord_polar`)

The Section 3 heatmap reads time on a Cartesian axis. Wrapping the hour-of-day totals around `coord_polar()` turns them into a clock face, making the daily rhythm of attacks immediately legible.

```{r}
hourly <- attacks_cleaned %>%
  count(hour) %>%
  na.omit()

p_clock <- ggplot(hourly, aes(x = hour, y = n, fill = n)) +
  geom_col(width = 1, colour = "white", linewidth = 0.2) +
  coord_polar(start = 0) +
  scale_fill_viridis(name = "# of attacks", option = "magma", direction = -1) +
  labs(title = "A 24-hour clock of attack volume",
       subtitle = "Each wedge is one hour of the day (local time)",
       x = NULL, y = NULL) +
  theme_minimal(base_family = "Helvetica") +
  theme(axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 8))

ggsave("images/extra_attack_clock.png", p_clock, width = 6, height = 6, dpi = 150)
p_clock
```

## GitHub-style calendar grid heatmap (`lubridate` + `viridis`)

The reference "calendar heatmap" is really a weekday-by-hour grid. Here I build a **true calendar layout**: one panel per month, weekdays across the columns and week-of-month down the rows, each day-cell shaded by its total attack count - exactly the layout GitHub uses for contribution graphs.

```{r}
daily <- attacks %>%
  mutate(date = as_date(ymd_hms(timestamp))) %>%
  count(date) %>%
  mutate(
    month         = month(date, label = TRUE, abbr = FALSE),
    wday          = wday(date, week_start = 1),                # 1 = Mon ... 7 = Sun
    first_wday    = wday(floor_date(date, "month"), week_start = 1),
    week_of_month = ceiling((day(date) + first_wday - 1) / 7)
  )

p_calendar <- ggplot(daily, aes(x = wday, y = week_of_month, fill = n)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  facet_wrap(~ month, ncol = 4) +
  scale_y_reverse(breaks = NULL) +
  scale_x_continuous(breaks = 1:7,
                     labels = c("M", "T", "W", "T", "F", "S", "S")) +
  scale_fill_viridis(name = "attacks/day", option = "viridis", direction = -1) +
  coord_equal() +
  labs(title = "Daily attack counts laid out as a calendar (2015)",
       x = NULL, y = NULL) +
  theme_minimal(base_family = "Helvetica") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold"))

ggsave("images/extra_calendar_grid.png", p_calendar, width = 9, height = 5, dpi = 150)
p_calendar
```

## Side-by-side cycle plots (`gridExtra::grid.arrange`)

`gridExtra` is loaded by both reference files but never used. I put it to work to stack two cycle plots into a single comparison panel - here South Korea vs China arrivals from 2010 - so their seasonal shapes can be read against each other.

```{r}
cycle_plot <- function(df, country) {
  d <- df %>%
    transmute(month, year, value = .data[[country]]) %>%
    filter(year >= 2010)
  hl <- d %>% group_by(month) %>% summarise(avgvalue = mean(value))

  ggplot() +
    geom_line(data = d, aes(year, value, group = month),
              colour = "grey25", linewidth = 0.4) +
    geom_hline(data = hl, aes(yintercept = avgvalue),
               colour = "firebrick", linetype = "dashed", linewidth = 0.4) +
    facet_grid(~ month) +
    scale_y_continuous(labels = label_comma()) +
    labs(title = country, x = NULL, y = "Visitors") +
    theme_tufte(base_family = "Helvetica") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          plot.title = element_text(hjust = 0.5, size = 11),
          strip.text = element_text(size = 7))
}

p_kr <- cycle_plot(air, "South Korea")
p_cn <- cycle_plot(air, "China")

p_compare <- arrangeGrob(p_kr, p_cn, ncol = 1,
                         top = "Cycle plots compared: South Korea vs China, 2010-2019")
ggsave("images/extra_cycle_compare.png", p_compare, width = 10, height = 6, dpi = 150)
grid.arrange(p_compare)
```

## Rice-yield growth ranking (`scales::percent`)

The slopegraph shows two endpoints; this extra distils the whole story into a ranking. I compute each country's **percentage change in rice yield from 1961 to 2010** and draw a lollipop chart sorted by growth, labelled with `scales::percent()`.

```{r}
rice_growth <- rice %>%
  filter(Year %in% c(1961, 2010)) %>%
  select(Country, Year, Yield) %>%
  pivot_wider(names_from = Year, values_from = Yield, names_prefix = "y") %>%
  mutate(growth = (y2010 - y1961) / y1961) %>%
  arrange(growth) %>%
  mutate(Country = factor(Country, levels = Country))

p_growth <- ggplot(rice_growth, aes(x = growth, y = Country)) +
  geom_segment(aes(x = 0, xend = growth, y = Country, yend = Country),
               colour = "grey75") +
  geom_point(aes(colour = growth), size = 4) +
  geom_text(aes(label = percent(growth, accuracy = 1)),
            hjust = -0.25, size = 3) +
  scale_x_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0.02, 0.18))) +
  scale_colour_viridis(option = "D", guide = "none") +
  labs(title = "Growth in rice yield, 1961 to 2010",
       subtitle = "Percentage change in yield (kg/ha) per country",
       x = NULL, y = NULL) +
  theme_minimal(base_family = "Helvetica") +
  theme(panel.grid.major.y = element_blank(),
        plot.subtitle = element_text(size = 8))

ggsave("images/extra_yield_growth.png", p_growth, width = 8, height = 5, dpi = 150)
p_growth
```

# 7 References

-   Kam, T.S. *R for Visual Analytics* - [Chapter 17: Visualising and Analysing Time-oriented Data](https://r4va.netlify.app/chap17).
-   [`ggHoriPlot`](https://rivasiker.github.io/ggHoriPlot/) - horizon plots in `ggplot2`.
-   [`CGPfunctions`](https://ibecav.github.io/CGPfunctions/) - `newggslopegraph()` for slopegraphs.
-   [`viridis`](https://sjmgarnier.github.io/viridis/) and [`scales`](https://scales.r-lib.org/) - colour scales and axis formatting.