---
title: "Hands-on Exercise 7B:"
subtitle: "Visualising and Analysing Text Data with R — tidytext methods"
date: "2 Jun 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1 Overview

This hands-on exercise covers [Chapter 29: Visualising and Analysing Text Data with R: tidytext methods](https://r4va.netlify.app/chap29).

By the end of this exercise, I will be able to:

-   understand the tidytext framework for processing, analysing and visualising text data,

-   write a function for importing multiple files into R,

-   combine multiple files into a single data frame,

-   clean and wrangle text data using the tidyverse approach,

-   visualise words with a Word Cloud,

-   compute term frequency–inverse document frequency (TF-IDF) using tidytext, and

-   visualise the relationships between texts and terms.

# 2 Getting Started

## 2.1 Loading the required packages

For this exercise we will use the following R packages:

-   **tidytext** and **tidyverse** (mainly readr, purrr, stringr, ggplot2) for the tidy text workflow,

-   **widyr** for pairwise counts/correlations,

-   **wordcloud** and **ggwordcloud** for word clouds,

-   **textplot** (which depends on **igraph**, **tidygraph** and **ggraph**) for term networks,

-   **DT** for interactive tables, and

-   **lubridate** and **hms** for date/time handling.

```{r}
#| label: setup
pacman::p_load(tidytext, widyr, wordcloud, DT, ggwordcloud,
               textplot, lubridate, hms, tidyverse,
               tidygraph, ggraph, igraph)

# Set seed so the stochastic layouts/word clouds are reproducible
set.seed(1234)
```

## 2.2 Importing Multiple Text Files from Multiple Folders

### 2.2.1 Creating a folder list

```{r}
news20 <- "data/20news"
```

### 2.2.2 Define a function to read all files from a folder into a data frame

```{r}
read_folder <- function(infolder) {
  tibble(file = dir(infolder,
                    full.names = TRUE)) %>%
    mutate(text = map(file,
                      read_lines)) %>%
    transmute(id = basename(file),
              text) %>%
    unnest(text)
}
```

## 2.3 Importing Multiple Text Files from Multiple Folders

### 2.3.1 Reading in all the messages from the 20news folder

```{r}
raw_text <- tibble(folder =
                     dir(news20,
                         full.names = TRUE)) %>%
  mutate(folder_out = map(folder,
                          read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  transmute(newsgroup = basename(folder),
            id, text)

# Make sure the output folder exists before writing the cache
if (!dir.exists("data/rds")) dir.create("data/rds", recursive = TRUE)
write_rds(raw_text, "data/rds/news20.rds")
```

# 3 Initial EDA

We can visualise the frequency of messages by newsgroup.

```{r}
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```

::: {.callout-tip appearance="simple"}
For each newsgroup, there are 10 different ids (messages).
:::

# 4 Introducing tidytext

-   Using tidy data principles in processing, analysing and visualising text data.

-   Much of the infrastructure needed for text mining with tidy data frames already exists in packages like *dplyr*, *broom*, *tidyr*, and *ggplot2*.

The workflow using the tidytext approach is: **read text → unnest tokens → remove stop words → summarise / visualise**.

## 4.1 Removing header and automated email signatures

Each message has some structure and extra text we don't want to include. Every message has a header (fields such as "from:" or "in_reply_to:"), and some have automated email signatures that occur after a line like "--".

```{r}
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

::: {.callout-note appearance="simple"}
This code chunk removes the non-message text from the raw text.
:::

## 4.2 Removing lines with nested text representing quotes from other users

```{r}
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text,
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text,
                     "^In article <")
  )
```

::: {.callout-tip appearance="simple"}
This removed quotes from other users so they are not double counted / analysed.
:::

## 4.3 Text Data Processing

[`unnest_tokens()`](https://www.rdocumentation.org/packages/tidytext/topics/unnest_tokens) of **tidytext** is used to split the dataset into tokens, while `stop_words` is used to remove stop words.

```{r}
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)
```

Now that we've removed the headers, signatures, and formatting, we can explore the most common words.

::: panel-tabset
### Most common words overall

```{r}
usenet_words %>%
  count(word, sort = TRUE) %>%
  datatable()
```

### Most common words by newsgroup

```{r}
words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()

words_by_newsgroup %>% datatable()
```
:::

## 4.4 Visualising Words in newsgroups using the wordcloud package

We can add colours for convenience using palettes from **RColorBrewer**.

```{r}
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300,
          colors = brewer.pal(9, "Dark2"))
```

A DT table can complement the visual discovery. We specify `filter` to add a text filter per column.

```{r}
datatable(words_by_newsgroup, filter = "top")
```

## 4.5 Visualising Words in newsgroups using the ggwordcloud package

This can be used along with **ggplot2**. Below we render facets for each newsgroup.

**Note:** We use `n > 8` so the word cloud renders faster and the words do not overlap.

```{r}
#| fig-width: 11
#| fig-height: 9
words_by_newsgroup %>%
  filter(n > 8) %>%
  ggplot(aes(label = word,
             size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

# 5 Basic Concept of TF-IDF

[tf–idf](https://en.wikipedia.org/wiki/Tf%E2%80%93idf) — term frequency–inverse document frequency — is a numerical statistic intended to reflect how important a word is to a document within a [corpus](https://en.wikipedia.org/wiki/Text_corpus).

$$tf\text{-}idf(t, d) = tf(t, d) \times \ln\left(\frac{N}{n_t}\right)$$

where $tf(t,d)$ is the frequency of term $t$ in document $d$, $N$ is the number of documents, and $n_t$ is the number of documents containing $t$.

## 5.1 Computing tf-idf within newsgroups

`bind_tf_idf()` of **tidytext** computes and binds the term frequency, inverse document frequency and tf-idf of a tidy text dataset.

```{r}
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

## 5.2 Visualising tf-idf as an interactive table

We use `formatRound()` to format the numeric columns and `formatStyle()` to add styling.

```{r}
datatable(tf_idf, filter = "top") %>%
  formatRound(c("tf", "idf", "tf_idf"), 3) %>%
  formatStyle(0,
              target = 'row',
              lineHeight = '25%')
```

## 5.3 Visualising tf-idf within newsgroups

A faceted bar chart is used to visualise the tf-idf values of the science-related newsgroups.

```{r}
#| fig-width: 10
#| fig-height: 7
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf,
            n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word,
                        tf_idf)) %>%
  ggplot(aes(tf_idf,
             word,
             fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup,
             scales = "free") +
  labs(x = "tf-idf",
       y = NULL)
```

# 6 Word Correlations

## 6.1 Counting and correlating pairs of words with the widyr package

-   To count the number of times that two words appear within the same document, or to see how correlated they are.

-   Most operations for finding pairwise counts or correlations need to turn the data into a wide matrix first.

-   The [**widyr**](https://cran.r-project.org/web/packages/widyr/) package first 'casts' a tidy dataset into a wide matrix, performs an operation such as a correlation on it, then re-tidies the result.

`pairwise_cor()` of **widyr** computes the correlation between newsgroups based on their common words.

```{r}
newsgroup_cors <- words_by_newsgroup %>%
  pairwise_cor(newsgroup,
               word,
               n,
               sort = TRUE)
```

## 6.2 Visualising correlation as a network

We can visualise the relationship between newsgroups in a network graph.

```{r}
set.seed(1234)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation,
                     width = correlation)) +
  geom_node_point(size = 6,
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 color = "red",
                 repel = TRUE) +
  theme_minimal()
```

::: {.callout-note appearance="simple"}
Result differs from the original chapter as we are using seed `1234` instead of `2017`.
:::

## 6.3 Bigram

A bigram data frame is created by using `unnest_tokens()` of tidytext with `token = "ngrams"` and `n = 2`.

```{r}
bigrams <- cleaned_text %>%
  unnest_tokens(bigram,
                text,
                token = "ngrams",
                n = 2)
bigrams
```

## 6.4 Counting bigrams

```{r}
bigrams_count <- bigrams %>%
  filter(bigram != 'NA') %>%
  count(bigram, sort = TRUE)
bigrams_count
```

## 6.5 Cleaning bigram

The code chunk below separates the bigram into two words and removes stop words.

```{r}
bigrams_separated <- bigrams %>%
  filter(bigram != 'NA') %>%
  separate(bigram, c("word1", "word2"),
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)

bigrams_filtered
```

## 6.6 Counting the bigram again

After filtering out the bigrams with stop words, we recount.

```{r}
bigram_counts <- bigrams_filtered %>%
  count(word1, word2, sort = TRUE)
bigram_counts
```

## 6.7 Create a network graph from the bigram data frame

A network graph is created using `graph_from_data_frame()` of **igraph**.

```{r}
bigram_graph <- bigram_counts %>%
  filter(n > 3) %>%
  graph_from_data_frame()
bigram_graph
```

## 6.8 Visualising a network of bigrams with ggraph

```{r}
set.seed(1234)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1)
```

## 6.9 Revised version

We can improve the visualisation by adding colour and directional arrows.

```{r}
set.seed(1234)

a <- grid::arrow(type = "closed",
                 length = unit(.15,
                               "inches"))

ggraph(bigram_graph,
       layout = "fr") +
  geom_edge_link(aes(edge_alpha = n),
                 show.legend = FALSE,
                 arrow = a,
                 end_cap = circle(.02,
                                  'inches')) +
  geom_node_point(color = "lightblue",
                  size = 2) +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1) +
  theme_minimal()
```

# 7 Extras

Beyond the guided exercise, I explored five additional text-analytics ideas. They reuse the `usenet_words`, `words_by_newsgroup` and `bigrams_filtered` objects built above.

## 7.1 Sentiment analysis with the Bing lexicon

I join the tokens with the **Bing** sentiment lexicon (bundled with **tidytext** as the `sentiments` dataset, so no interactive download is needed) and chart the words that contribute most to positive and negative sentiment.

```{r}
# Robustly obtain the Bing lexicon across tidytext versions, offline
bing_lex <- sentiments
if ("lexicon" %in% names(bing_lex)) {
  bing_lex <- dplyr::filter(bing_lex, lexicon == "bing")
}
bing_lex <- bing_lex %>%
  select(word, sentiment) %>%
  filter(sentiment %in% c("positive", "negative"))

word_sentiments <- usenet_words %>%
  inner_join(bing_lex, by = "word") %>%
  count(word, sentiment, sort = TRUE)

word_sentiments %>%
  group_by(sentiment) %>%
  slice_max(n, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(title = "Words contributing most to sentiment",
       x = "Contribution to sentiment", y = NULL) +
  theme_minimal()
```

## 7.2 Which newsgroups are most positive or negative?

Aggregating the same lexicon by newsgroup gives a **net sentiment** score (positive minus negative word counts) per group.

```{r}
newsgroup_sentiment <- usenet_words %>%
  inner_join(bing_lex, by = "word") %>%
  count(newsgroup, sentiment) %>%
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>%
  mutate(net_sentiment = positive - negative) %>%
  arrange(net_sentiment)

ggplot(newsgroup_sentiment,
       aes(x = reorder(newsgroup, net_sentiment),
           y = net_sentiment,
           fill = net_sentiment > 0)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#d73027", "TRUE" = "#1a9850")) +
  labs(title = "Net sentiment by newsgroup (Bing lexicon)",
       x = NULL, y = "Net sentiment (positive − negative)") +
  theme_minimal()
```

::: {.callout-note appearance="simple"}
Newsgroups such as `talk.politics.*` tend toward negative net sentiment, while `misc.forsale` and the sports groups skew positive — a sanity check that the pipeline captures meaningful signal.
:::

## 7.3 Word-correlation network *within* a single newsgroup

In §6 we correlated whole newsgroups. Here I zoom into **sci.space** and use `pairwise_cor()` on *words* across messages (`id`) to find which terms tend to co-occur, then draw the correlation network.

```{r}
set.seed(1234)

space_word_cors <- usenet_words %>%
  filter(newsgroup == "sci.space") %>%
  group_by(word) %>%
  filter(n() >= 15) %>%      # keep reasonably frequent words
  ungroup() %>%
  pairwise_cor(word, id, sort = TRUE)

space_word_cors %>%
  filter(correlation > 0.15) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation, width = correlation),
                 show.legend = FALSE) +
  geom_node_point(size = 5, color = "lightblue") +
  geom_node_text(aes(label = name), repel = TRUE) +
  labs(title = "Word correlations within sci.space") +
  theme_minimal()
```

The strongest pairs as a table:

```{r}
space_word_cors %>%
  slice_max(correlation, n = 15) %>%
  datatable() %>%
  formatRound("correlation", 3)
```

## 7.4 Most distinctive *bigrams* with TF-IDF

TF-IDF is not limited to single words. By computing TF-IDF on **bigrams**, we surface the two-word phrases that most distinguish each science newsgroup.

```{r}
#| fig-width: 10
#| fig-height: 7
bigram_tf_idf <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(newsgroup, bigram, sort = TRUE) %>%
  bind_tf_idf(bigram, newsgroup, n) %>%
  arrange(desc(tf_idf))

bigram_tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(bigram = reorder_within(bigram, tf_idf, newsgroup)) %>%
  ggplot(aes(tf_idf, bigram, fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~newsgroup, scales = "free") +
  scale_y_reordered() +
  labs(title = "Most distinctive bigrams (TF-IDF) in sci.* newsgroups",
       x = "tf-idf", y = NULL)
```

## 7.5 A reusable per-newsgroup word-cloud function

Finally I wrap the word-cloud logic in a small reusable function, so any newsgroup can be visualised on demand. I demonstrate it on two contrasting groups in a tabset.

```{r}
plot_newsgroup_cloud <- function(ng, max_words = 80) {
  dat <- words_by_newsgroup %>% filter(newsgroup == ng)
  wordcloud(dat$word, dat$n,
            max.words = max_words,
            scale = c(3, 0.4),
            colors = brewer.pal(8, "Dark2"))
}
```

::: panel-tabset
### sci.space

```{r}
plot_newsgroup_cloud("sci.space")
```

### rec.sport.hockey

```{r}
plot_newsgroup_cloud("rec.sport.hockey")
```
:::

::: {.callout-tip appearance="simple"}
Because the function returns to the standard `wordcloud()` call, the same helper can be dropped into any report and pointed at any of the 20 newsgroups.
:::

# 8 References

-   Kam, T.S. (2024). [Visualising and Analysing Text Data with R: tidytext methods](https://r4va.netlify.app/chap29). *R for Visual Analytics.*

-   Silge, J. & Robinson, D. *Text Mining with R: A Tidy Approach.* <https://www.tidytextmining.com/>
