---
title: "Hands-on Exercise 9C:"
subtitle: "Analytical Mapping"
date: "19 Jun 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1 Overview

This hands-on exercise covers [Chapter 23: Analytical Mapping](https://r4va.netlify.app/chap23).

Analytical mapping goes beyond pretty choropleths toward **spatial exploratory
data analysis (ESDA)**. In this exercise I:

-   Import an `rds` polygon layer of Nigeria water points by LGA.
-   Map **rates** (proportions) rather than raw counts.
-   Build **percentile maps** and **box maps** to surface spatial outliers.
-   Write reusable functions (`get.var`, `percentmap`, `boxbreaks`, `boxmap`).

# 2 Getting Started

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

`NGA_wp.rds` is a polygon feature data frame of water-point information across
Nigeria's Local Government Areas (LGAs).

```{r}
NGA_wp <- read_rds("data/NGA_wp.rds")
glimpse(NGA_wp)
```

# 3 Basic Choropleth Mapping

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional", n = 10, style = "equal", palette = "Blues") +
  tm_borders(lwd = 0.1, alpha = 1) +
  tm_title("Functional water points by LGA", size = 0.9) +
  tm_layout(legend.outside = TRUE)

p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp", n = 10, style = "equal", palette = "Blues") +
  tm_borders(lwd = 0.1, alpha = 1) +
  tm_title("Total water points by LGA", size = 0.9) +
  tm_layout(legend.outside = TRUE)

tmap_arrange(p2, p1, nrow = 1)
```

# 4 Choropleth Map for Rates

Water points are **not equally distributed in space**, so mapping raw counts
mostly maps "where there are many water points" rather than the variable of
interest. Mapping **rates** corrects for that.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional / total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional / total_wp)
```

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional", n = 10, style = "equal",
          palette = "Blues", legend.hist = TRUE) +
  tm_borders(lwd = 0.1, alpha = 1) +
  tm_title("Rate map of functional water points by LGA", size = 0.9) +
  tm_layout(legend.outside = TRUE)
```

# 5 Extreme Value Maps

## Percentile map

Special quantile map with six categories: 0–1%, 1–10%, 10–50%, 50–90%, 90–99%,
99–100%.

```{r}
NGA_wp <- NGA_wp %>% drop_na()

get.var <- function(vname, df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}

percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam, title = legtitle, breaks = bperc, palette = "Blues",
            labels = c("< 1%", "1% - 10%", "10% - 50%",
                       "50% - 90%", "90% - 99%", "> 99%")) +
    tm_borders() +
    tm_title(mtitle) +
    tm_layout(legend.outside = TRUE)
}

percentmap("total_wp", NGA_wp)
```

## Box map

A box map is an augmented quartile map with extra lower/upper outlier classes.

```{r}
ggplot(data = NGA_wp, aes(x = "", y = wp_nonfunctional)) +
  geom_boxplot()
```

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  bb <- vector(mode = "numeric", length = 7)
  if (lofence < qv[1]) {
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) {
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}

boxmap <- function(vnam, df, legtitle = NA, mtitle = "Box Map", mult = 1.5) {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam, title = legtitle, breaks = bb, palette = "Blues",
            labels = c("lower outlier", "< 25%", "25% - 50%",
                       "50% - 75%", "> 75%", "upper outlier")) +
    tm_borders() +
    tm_title(mtitle) +
    tm_layout(legend.outside = TRUE)
}

tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

# 6 Extras {#sec-extras}

## Counts vs rates, side by side (proving the point)

The chapter *asserts* that counts mislead. Here I show it: the same LGAs that
look "worst" by raw non-functional count are different from those worst by
non-functional *rate*.

```{r}
count_map <- boxmap("wp_nonfunctional", NGA_wp,
                    mtitle = "Box map: non-functional COUNT")
rate_map  <- boxmap("pct_nonfunctional", NGA_wp,
                    mtitle = "Box map: non-functional RATE")
tmap_arrange(count_map, rate_map, ncol = 2)
```

::: callout-tip
## Insight
Large, populous LGAs dominate the **count** map simply because they have more of
everything. The **rate** map re-ranks small LGAs where most water points are
broken, which are the genuine service-failure hotspots.
:::

## Generalising percentile/box maps over many variables

The functions take the variable name as a string, so I can loop them across a
vector of variables instead of copy-pasting.

```{r}
#| fig-height: 5
vars <- c("pct_functional", "pct_nonfunctional")
box_maps <- lapply(vars, function(v) boxmap(v, NGA_wp, mtitle = v))
tmap_arrange(box_maps[[1]], box_maps[[2]], ncol = 2)
```

## Standard deviation map (another extreme-value lens)

A z-score or standard-deviation classification flags how far each LGA sits from
the mean, which is a different way of defining "extreme".

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_nonfunctional",
          style = "sd",
          palette = "-RdBu",
          midpoint = mean(NGA_wp$pct_nonfunctional, na.rm = TRUE),
          title = "Non-functional rate (std. dev.)") +
  tm_borders(lwd = 0.1, alpha = 0.6) +
  tm_title("Standard deviation map", size = 0.9) +
  tm_layout(legend.outside = TRUE)
```

## A raw-breaks helper to complete the function family

The `boxbreaks` function handles quartiles, so I add a small `rawbreaks`
companion that builds equal-count (quantile) breaks for any `n`, making the
function family self-contained.

```{r}
rawbreaks <- function(v, n = 6) {
  unname(quantile(v, probs = seq(0, 1, length.out = n + 1)))
}

ratemap <- function(vnam, df, n = 6, mtitle = "Rate Map") {
  var <- get.var(vnam, df)
  bb <- rawbreaks(var, n)
  tm_shape(df) +
    tm_fill(vnam, breaks = bb, palette = "Blues") +
    tm_borders(lwd = 0.1, alpha = 0.6) +
    tm_title(mtitle, size = 0.9) +
    tm_layout(legend.outside = TRUE)
}

ratemap("pct_functional", NGA_wp, n = 6, mtitle = "Functional rate (quantile breaks)")
```

## Interactive box map for exploration

Wrapping the box-map logic in `view` mode lets a reader click any LGA to read
its actual figures, pairing the classification with the raw evidence.

```{r}
tmap_mode("view")
NGA_valid <- st_make_valid(NGA_wp)
bb <- boxbreaks(get.var("wp_nonfunctional", NGA_valid))

tm_shape(NGA_valid) +
  tm_fill("wp_nonfunctional",
          breaks = bb,
          palette = "Blues",
          alpha = 0.75,
          id = "ADM2_EN",
          popup.vars = c("Total" = "total_wp",
                         "Functional" = "wp_functional",
                         "Non-functional" = "wp_nonfunctional",
                         "% non-functional" = "pct_nonfunctional"),
          labels = c("lower outlier", "< 25%", "25% - 50%",
                     "50% - 75%", "> 75%", "upper outlier")) +
  tm_borders(lwd = 0.5, alpha = 0.5)
tmap_mode("plot")
```

::: callout-note
The `id`/`popup.vars` field names (here `ADM2_EN`, the LGA name) depend on the
columns in `NGA_wp`; check `names(NGA_wp)` and adjust if your rds differs.
:::

# 7 Reflections

-   Analytical mapping is **EDA with a map attached**. Box and percentile maps
    are just boxplots and quantiles drawn in space, and that reframing made the
    custom functions click for me.
-   Extra 1 is the most important takeaway: **always look at counts and rates
    together** before drawing conclusions about "the worst" areas.
-   Writing the helpers as string-keyed functions (Extras 2 & 4) made it trivial
    to fan the same analysis across many variables.