---
title: "Hands-on Exercise 9B:" 
subtitle: "Visualising Geospatial Point Data"
date: "19 Jun 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1 Overview

This hands-on exercise covers [Chapter 22: Visualising Geospatial Point Data](https://r4va.netlify.app/chap22).

A **proportional symbol map** plots point locations and encodes a quantity in the
size (and/or colour) of the symbol. In this exercise I:

-   Import an **aspatial** csv and convert it into an `sf` point layer with the
    correct projection.
-   Draw interactive **proportional symbol maps** with `tmap`.
-   Encode a second variable with colour and use **synchronised facets**.

# 2 Getting Started

## Loading the required packages

-   **sf**: geospatial data handling
-   **tmap**: thematic mapping
-   **tidyverse**: data wrangling

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing data

The dataset is `SGPools_svy21.csv`: a list of Singapore Pools outlets with their
coordinates (`XCOORD`, `YCOORD`) in SVY21.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
list(rows = nrow(sgpools), cols = names(sgpools))
head(sgpools)
```

# 3 Data Wrangling

Convert the data frame into an `sf` object. `crs = 3414` is [EPSG:3414](https://epsg.io/3414),
Singapore's SVY21 projected coordinate system.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                       coords = c("XCOORD", "YCOORD"),
                       crs = 3414)
head(sgpools_sf)
```

::: callout-note
## Insight
The two coordinate columns are now collapsed into one `geometry` list-column of
type `POINT`. Getting the CRS right matters: if I had used a geographic CRS
(e.g. 4326) on SVY21 metre coordinates, every point would land in the wrong
place.
:::

# 4 Drawing a Proportional Symbol Map

Turn on interactive mode:

```{r}
tmap_mode("view")
```

## Plain interactive point map

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "red",
             size = 1,
             border.col = "black",
             border.lwd = 1)
```

## Proportional sizes

Map the bubble **size** to winnings:

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "red",
             size = "Gp1Gp2 Winnings",
             border.col = "black",
             border.lwd = 1)
```

## Adding colour

Encode `OUTLET TYPE` with colour:

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             border.col = "black",
             border.lwd = 1)
```

## Synchronised facets

`tm_facets(sync = TRUE)` links zoom/pan across the small multiples:

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             border.col = "black",
             border.lwd = 1) +
  tm_facets(by = "OUTLET TYPE", nrow = 1, sync = TRUE)
```

# 5 Extras 

## Adding spatial context (boundary basemap)

A floating cloud of bubbles is hard to read without a reference. I overlay the
outlets on the Singapore subzone boundary so location becomes meaningful.

```{r}
mpsz <- st_read(dsn = "data/MP14_SUBZONE_WEB_PL", layer = "MP14_SUBZONE_WEB_PL") %>%
  st_make_valid() %>%
  st_transform(3414)

tmap_mode("plot")
tm_shape(mpsz) +
  tm_polygons(col = "grey95", border.col = "white") +
  tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             border.col = "black",
             border.lwd = 0.5,
             alpha = 0.8) +
  tm_title("SG Pools outlets over planning subzones", size = 1) +
  tm_layout(legend.outside = TRUE)
```

## Choosing basemaps and richer popups

In `view` mode I can pick a basemap provider and control the hover tooltip,
turning the map into a small exploratory tool.

```{r}
tmap_mode("view")
tm_basemap(c("CartoDB.Positron", "OpenStreetMap", "Esri.WorldImagery")) +
  tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             id = "NAME",
             popup.vars = c("Type" = "OUTLET TYPE",
                            "Gp1/Gp2 Winnings" = "Gp1Gp2 Winnings",
                            "Address" = "ADDRESS"))
tmap_mode("plot")
```

## Aggregating points to areas (spatial join)

Individual points answer "where", but decision-makers often want "how many per
area". I spatially join outlets to planning areas and map the **count** as a
choropleth.

```{r}
outlets_by_area <- mpsz %>%
  st_join(sgpools_sf, join = st_intersects) %>%
  group_by(PLN_AREA_N) %>%
  summarise(n_outlets = sum(!is.na(`OUTLET TYPE`))) %>%
  ungroup()

tm_shape(outlets_by_area) +
  tm_fill("n_outlets",
          style = "jenks",
          palette = "Greens",
          title = "Outlets per planning area") +
  tm_borders(alpha = 0.4) +
  tm_title("Outlet density by planning area", size = 1) +
  tm_layout(legend.outside = TRUE)
```

## Total winnings per region

I aggregate the numeric winnings (not just counts) to compare regions.

```{r}
winnings_by_region <- mpsz %>%
  st_join(sgpools_sf, join = st_intersects) %>%
  st_drop_geometry() %>%
  group_by(REGION_N) %>%
  summarise(total_winnings = sum(`Gp1Gp2 Winnings`, na.rm = TRUE),
            outlets = sum(!is.na(`OUTLET TYPE`))) %>%
  arrange(desc(total_winnings))

winnings_by_region
```

## Reproducible static proportional symbol map

Interactive maps need a browser; a static version is needed for PDFs and print.
This `plot`-mode version reproduces the proportional symbol map and is fully
reproducible in a rendered document.

```{r}
tmap_mode("plot")
tm_shape(mpsz) +
  tm_polygons(col = "grey97", border.col = "grey85") +
  tm_shape(sgpools_sf) +
  tm_symbols(size = "Gp1Gp2 Winnings",
             col = "OUTLET TYPE",
             palette = c("#1b9e77", "#d95f02"),
             border.col = "black",
             border.lwd = 0.4,
             scale = 1.5,
             alpha = 0.75) +
  tm_title("SG Pools winnings (static proportional symbols)", size = 1) +
  tm_layout(legend.outside = TRUE, frame = FALSE)
```

# 6 Reflections

-   Points alone undersell the story; **context (Extra 1) and aggregation
    (Extras 3 to 4)** are what turn a scatter of bubbles into an insight.
-   Interactive maps are great for exploration, but keeping a **static fallback
    (Extra 5)** makes the analysis portable to any output format.
-   The recurring `st_make_valid()` step from Ex09A reappears here when joining
    against the subzone polygons, because invalid geometry is a constant tax on
    Singapore shapefiles.
