---
title: "Hands-on Exercise 9A:"
subtitle: "Choropleth Mapping with R"
date: "19 Jun 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

# 1 Overview

This hands-on exercise covers [Chapter 21: Choropleth Mapping with R](https://r4va.netlify.app/chap21).

A **choropleth map** is a thematic map where geographic areas (subzones, provinces, countries...) are shaded in proportion to a statistical variable. In this exercise I:

-   Import geospatial (shapefile) and aspatial (csv) data and join them.
-   Build choropleth maps with `tmap`, using `qtm()` and the layered `tm_*` element approach.
-   Compare data **classification methods** and **colour schemes**.
-   Customise **map layouts** (legends, compass, scale bar, grid, credits).
-   Produce **small multiples** and **criterion-based** maps.

# 2 Getting Started

## The data

Two datasets are used:

| Type | Dataset | Source |
|------|---------|--------|
| Geospatial | Master Plan 2014 Subzone Boundary (Web), `MP14_SUBZONE_WEB_PL` | [data.gov.sg](https://data.gov.sg/) |
| Aspatial | Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, 2011 to 2020 (`respopagesextod2011to2020.csv`) | [SingStat](https://www.singstat.gov.sg/) |

Expected folder structure:

``` bash
HO_EX09
└── data
    ├── respopagesextod2011to2020.csv
    ├── MP14_SUBZONE_WEB_PL
    │   └── MP14_SUBZONE_WEB_PL.{shp,shx,dbf,prj,...}
    └── rds          # created on first write
```

## Installing and loading R packages

```{r}
pacman::p_load(sf, tidyverse, tmap)
```

# 3 Data Wrangling

## Importing geospatial data

```{r}
mpsz <- st_read(dsn = "data/MP14_SUBZONE_WEB_PL", layer = "MP14_SUBZONE_WEB_PL")
```

## Importing aspatial data

The csv is aspatial, so `read_csv()` is used instead of `st_read()`:

```{r}
popdata <- read_csv("data/respopagesextod2011to2020.csv")
```

## Data preparation

I aggregate the 2020 population by planning area / subzone and derive the
demographic groups and the **dependency ratio**:

$$\text{Dependency} = \frac{\text{YOUNG} + \text{AGED}}{\text{ECONOMY ACTIVE}}$$

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

::: callout-note
## My read of the wrangling
The age-group columns are selected by position (`.[3:6]` and so on), which is
fragile. If the import order changes, the sums break. In the extras I keep the
same logic for fidelity, but in production I would select the age columns by name.
:::

The `SZ`/`PA` keys must be upper-cased to match `mpsz$SUBZONE_N` before joining:

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
#| eval: false
dir.create("data/rds", showWarnings = FALSE)
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

# 4 Choropleth Mapping with tmap

## Using `qtm()`

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

## Using tmap elements

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          title = "Dependency ratio") +
  tm_borders(alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from URA\n and Population data from DOS",
             position = c("left", "bottom")) +
  tm_title("Distribution of Dependency Ratio by planning subzone", size = 1.1) +
  tm_layout(legend.outside = TRUE, frame = TRUE)
```

## Classification methods

### Quantile vs Equal

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY", n = 5, style = "quantile") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY", n = 5, style = "equal") +
  tm_borders(alpha = 0.5)
```

::: callout-tip
## Insight
There is a strong outlier subzone in the dependency ratio. With **equal
interval** classes, almost every subzone falls in the first bucket and the map
looks flat. **Quantile** classification guarantees roughly equal-sized groups,
so spatial variation is visible despite the outlier.
:::

## Colour scheme

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY", n = 6, style = "quantile", palette = "Blues") +
  tm_borders(alpha = 0.5)
```

## Map layout

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "jenks",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio (Jenks)", size = 1) +
  tm_layout(legend.outside = TRUE, frame = FALSE)
```

## Small multiples and criterion mapping

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_borders(alpha = 0.5) +
  tm_layout(legend.outside = TRUE)
```

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Greens",
          legend.hist = TRUE) +
  tm_layout(legend.outside = TRUE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

# 5 Extras

## Fixing the interactive (`view`) map

Earlier, `tmap_mode("view")` failed with an *invalid polygon* error. The cause
is self-intersecting geometries in the shapefile, and `st_make_valid()` repairs
them so the map can render in Leaflet.

```{r}
mpsz_valid <- st_make_valid(mpsz_pop2020)

tmap_mode("view")
tm_shape(mpsz_valid) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          alpha = 0.7,
          id = "SUBZONE_N",
          popup.vars = c("Dependency" = "DEPENDENCY",
                         "Aged" = "AGED",
                         "Young" = "YOUNG")) +
  tm_borders(alpha = 0.5)
tmap_mode("plot")
```

::: callout-note
The interactive map lets a reader hover over any subzone to read its underlying
numbers, which a static choropleth cannot do. `popup.vars` controls what the
tooltip shows.
:::

## Comparing six classification methods at a glance

Instead of viewing methods one at a time, I render six in a single grid so the
visual differences are obvious.

```{r}
#| fig-height: 8
styles <- c("equal", "quantile", "jenks", "kmeans", "sd", "pretty")
maps <- lapply(styles, function(s) {
  tm_shape(mpsz_pop2020) +
    tm_fill("DEPENDENCY", style = s, n = 5, palette = "Blues",
            title = s) +
    tm_borders(alpha = 0.4) +
    tm_layout(main.title = s, main.title.size = 0.9, legend.show = FALSE)
})
tmap_arrange(maps[[1]], maps[[2]], maps[[3]],
             maps[[4]], maps[[5]], maps[[6]], ncol = 3)
```

::: callout-tip
`quantile`, `jenks` and `kmeans` reveal the most spatial structure; `equal` and
`sd` are dominated by the outlier and look nearly uniform.
:::

## Effect of the number of classes

Same variable and method, varying `n`. Too few classes hide variation; too many
overload the reader.

```{r}
#| fig-height: 6
n_list <- c(2, 5, 10, 20)
n_maps <- lapply(n_list, function(k) {
  tm_shape(mpsz_pop2020) +
    tm_fill("DEPENDENCY", style = "quantile", n = k, palette = "Blues") +
    tm_borders(alpha = 0.3) +
    tm_layout(main.title = paste0("n = ", k), main.title.size = 0.9,
              legend.show = FALSE)
})
tmap_arrange(n_maps[[1]], n_maps[[2]], n_maps[[3]], n_maps[[4]], ncol = 2)
```

## Custom (manual) class breaks

Classification can be driven by domain knowledge instead of an algorithm. Here I
set breaks from the distribution's own descriptive statistics.

```{r}
summary(mpsz_pop2020$DEPENDENCY)

breaks <- c(0, 0.6, 0.7, 0.8, 0.9, Inf)
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          breaks = breaks,
          palette = "Blues",
          title = "Dependency ratio (custom breaks)") +
  tm_borders(alpha = 0.5) +
  tm_layout(legend.outside = TRUE)
```

## Colour-blind-safe palette and ranked table

Choropleths must communicate to *all* readers. I switch to the colour-blind-safe
`viridis` palette and add a table of the 10 most dependent subzones so the map's
extremes are also legible as numbers.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = viridisLite::viridis(6),
          title = "Dependency (viridis)") +
  tm_borders(alpha = 0.4)
```

```{r}
mpsz_pop2020 %>%
  st_drop_geometry() %>%
  filter(!is.na(DEPENDENCY)) %>%
  arrange(desc(DEPENDENCY)) %>%
  select(SUBZONE_N, PLN_AREA_N, YOUNG, AGED, `ECONOMY ACTIVE`, DEPENDENCY) %>%
  slice_head(n = 10)
```

# 6 Reflections

-   The biggest practical lesson is that **classification choice is an editorial
    decision**, not a default. Extras 2 to 4 show the *same data* telling very
    different stories depending on style, `n`, and breaks.
-   `st_make_valid()` (Extra 1) is the small fix that unlocks the entire
    interactive workflow, and it is worth remembering for every shapefile-based map.
-   Accessibility (Extra 5) is easy to bolt on and should be the default, not an
    afterthought.
