---
title: "Hands-on Exercise 10:"
subtitle: "Information Dashboard Design with R"
date: "25 Jun 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  fig-width: 8.5
  fig-align: center
---

# 1 Overview

This hands-on exercise covers [Chapter 31: Information Dashboard Design: R methods](https://r4va.netlify.app/chap31).

An **information dashboard** packs the key numbers a decision-maker needs into a
single, scannable view. The design challenge is density without clutter: each
cell has to earn its space. In this exercise I:

-   Build the classic dashboard building blocks in `ggplot2`: the **bullet
    chart** (actual vs target) and the **sparkline** (compact trend lines).
-   Assemble a **static dashboard** with `gt` and `gtExtras`, embedding bullets
    and sparklines directly inside a table.
-   Assemble an **interactive dashboard** with `reactable` and `reactablefmtr`,
    adding hover, points, reference lines and bandlines.

The 5 extras (Section [-@sec-extras]) push the same toolkit further: KPI
scorecards, dual-metric sparklines, a cross-product sales heatmap, trend small
multiples, and an exception-reporting view.

# 2 Getting Started

## 2.1 Loading the required packages

-   **tidyverse** and **lubridate**: data wrangling and date handling
-   **gt** and **gtExtras**: publication-quality static tables with embedded plots
-   **reactable** and **reactablefmtr**: interactive HTML tables
-   **ggthemes**: extra `ggplot2` themes
-   **svglite**: SVG device needed by `gt_plt_bullet()`

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
               reactablefmtr, gt, gtExtras, tidyverse, svglite)
```

I view this site in dark mode, where `ggplot2`'s default grey text washes out
against the white plot background. I set a global theme so every figure uses
**black text on a white background**, and reuse the same overrides on the plots
that supply their own theme.

```{r}
black_text <- theme(
  text = element_text(colour = "black"),
  axis.text = element_text(colour = "black"),
  plot.title = element_text(colour = "black", face = "bold"),
  plot.background = element_rect(fill = "white", colour = NA),
  legend.background = element_rect(fill = "white", colour = NA)
)

theme_set(theme_gray(base_size = 13) + black_text)
```

The interactive sparklines in Section 5 rely on the **dataui** package, which
lives on GitHub rather than CRAN. I install it once and then load it:

```{r}
#| eval: false
remotes::install_github("timelyportfolio/dataui")
```

```{r}
library(dataui)
```

## 2.2 Importing data

The dataset is `CoffeeChain.rds`, a sales record of a fictional coffee retail
chain (4,248 rows). It carries actuals and budgets for `Sales`, `Profit`,
`Margin` and related measures, plus `Product`, `Product Type`, `Market` and a
daily `Date`.

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")
glimpse(coffeechain)
```

# 3 Plotting the Building Blocks with ggplot

## 3.1 Bullet chart

A bullet chart compares a single measure (the actual) against a target and a set
of qualitative ranges. I first aggregate `Sales` and `Budget Sales` to the
product level.

```{r}
product <- coffeechain %>%
  group_by(Product) %>%
  summarise(target = sum(`Budget Sales`),
            current = sum(Sales)) %>%
  ungroup()
```

The chart is built from layered `geom_col()` bars (the grey qualitative bands),
a thin black bar (the actual), and a red error bar standing in for the target
marker.

```{r}
#| fig-height: 6
ggplot(product, aes(Product, current)) +
  geom_col(aes(Product, max(target) * 1.01),
           fill = "grey85", width = 0.85) +
  geom_col(aes(Product, target * 0.75),
           fill = "grey60", width = 0.85) +
  geom_col(aes(Product, target * 0.5),
           fill = "grey50", width = 0.85) +
  geom_col(aes(Product, current),
           width = 0.35, fill = "black") +
  geom_errorbar(aes(y = target, x = Product, ymin = target, ymax = target),
                width = 0.4, colour = "red", linewidth = 1) +
  labs(x = "Product", y = "Total sales (actual vs budget)") +
  coord_flip()
```

## 3.2 Sparklines

### 3.2.1 Preparing the data

Sparklines need one small time series per product. I build a monthly sales
report, then derive the supporting points (minimum, maximum, month-end) and the
inner-quartile ribbon.

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

```{r}
mins <- group_by(sales_report, Product) %>%
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>%
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>%
  filter(Month == max(Month))
```

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 0.25),
            quart2 = quantile(Sales, 0.75)) %>%
  right_join(sales_report)
```

### 3.2.2 Generating the chart

```{r}
#| fig-height: 7
ggplot(sales_report, aes(x = Month, y = Sales)) +
  facet_grid(Product ~ ., scales = "free_y") +
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2),
              fill = "grey90") +
  geom_line(linewidth = 0.3) +
  geom_point(data = mins, col = "red") +
  geom_point(data = maxs, col = "blue") +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) +
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), strip.text = element_blank())
```

::: callout-note
## Insight
These two `ggplot` versions are the "from scratch" reference. They work, but the
sparkline in particular takes a lot of code for a small result. Sections 4 and 5
show how purpose-built table packages collapse the same output into a line or
two.
:::

# 4 Static Dashboard: gt and gtExtras

## 4.1 Bullet chart inside a table

`gt_plt_bullet()` reproduces the Section 3.1 chart from the aggregated `product`
table in a single call.

```{r}
product %>%
  gt() %>%
  gt_plt_bullet(column = current,
                target = target,
                width = 60,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

## 4.2 Sparklines with gtExtras

### 4.2.1 Preparing the data

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()

head(report)
```

`gt_plt_sparkline()` expects the values as a list-column, so I collapse each
product's monthly sales into a list.

```{r}
report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop")
```

### 4.2.2 Plotting sparklines

```{r}
report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop") %>%
  gt() %>%
  gt_plt_sparkline(`Monthly Sales`, same_limit = FALSE)
```

### 4.2.3 Adding summary statistics

```{r}
report %>%
  group_by(Product) %>%
  summarise(Min = min(Sales, na.rm = TRUE),
            Max = max(Sales, na.rm = TRUE),
            Average = mean(Sales, na.rm = TRUE)) %>%
  gt() %>%
  fmt_number(columns = 4, decimals = 2)
```

### 4.2.4 Combining sparklines, statistics and a bullet chart

I build the sparkline list, the summary statistics and the bullet inputs
separately, then join them into one table.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop")

sales <- report %>%
  group_by(Product) %>%
  summarise(Min = min(Sales, na.rm = TRUE),
            Max = max(Sales, na.rm = TRUE),
            Average = mean(Sales, na.rm = TRUE))

bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(Product) %>%
  summarise(Target = sum(`Budget Sales`),
            Actual = sum(Sales)) %>%
  ungroup()

sales_data <- sales %>%
  left_join(spark) %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline(`Monthly Sales`) %>%
  gt_plt_bullet(column = Actual,
                target = Target,
                width = 28,
                palette = c("lightblue", "black")) %>%
  gt_theme_538()
```

::: callout-tip
## Insight
This single table is already a dashboard: each row shows a product's trend
(sparkline), its spread (min/max/average) and its performance against budget
(bullet). It reads top to bottom like a management report.
:::

# 5 Interactive Dashboard: reactable and reactablefmtr

`reactable` renders sortable, searchable HTML tables, and `reactablefmtr`'s
`react_sparkline()` turns a list-column into an interactive sparkline. I first
reshape the report into one list per product.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

## 5.1 Interactive sparklines

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(cell = react_sparkline(report))
  )
)
```

## 5.2 Adding points and labels

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(min = "red", max = "blue"),
        labels = c("first", "last"))
    )
  )
)
```

## 5.3 Adding a reference line

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(min = "red", max = "blue"),
        statline = "mean")
    )
  )
)
```

## 5.4 Adding a bandline

A bandline shades the inner quartiles, making each product's spread visible at a
glance.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green")
    )
  )
)
```

## 5.5 Using sparkbars

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

# 6 Extras {#sec-extras}

## 6.1 A KPI scorecard by market

A dashboard usually opens with headline numbers. I summarise the chain by
`Market` and use `gt` with `data_color()` to turn the table into a heat-shaded
scorecard, so the strongest and weakest markets read instantly.

```{r}
market_kpi <- coffeechain %>%
  group_by(Market) %>%
  summarise(Sales = sum(Sales),
            Profit = sum(Profit),
            Margin = sum(Profit) / sum(Sales)) %>%
  arrange(desc(Sales))

market_kpi %>%
  gt() %>%
  tab_header(title = "Coffee Chain KPI Scorecard",
             subtitle = "Total performance by market") %>%
  fmt_currency(columns = c(Sales, Profit), currency = "USD", decimals = 0) %>%
  fmt_percent(columns = Margin, decimals = 1) %>%
  data_color(columns = Sales, palette = c("white", "#1b7837")) %>%
  data_color(columns = Profit, palette = c("white", "#2166ac")) %>%
  data_color(columns = Margin, palette = c("#b2182b", "white", "#1b7837")) %>%
  gt_theme_538()
```

::: callout-tip
## Insight
Colour does the ranking for the reader. A market can post high sales yet a thin
margin, and the diverging colour scale on `Margin` surfaces exactly that case,
which a plain number column would hide.
:::

## 6.2 Comparing two trends: sales vs profit sparklines

A single sales sparkline hides whether revenue and profit move together. I place
a sales sparkline and a profit sparkline side by side for each product, so
divergence between the two is obvious.

```{r}
dual_spark <- coffeechain %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  filter(year(Date) == 2013) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales), Profit = sum(Profit), .groups = "drop") %>%
  group_by(Product) %>%
  summarise(`Sales trend` = list(Sales),
            `Profit trend` = list(Profit), .groups = "drop")

dual_spark %>%
  gt() %>%
  gt_plt_sparkline(`Sales trend`, same_limit = FALSE) %>%
  gt_plt_sparkline(`Profit trend`, same_limit = FALSE) %>%
  gt_theme_538()
```

## 6.3 Sales heatmap by product and month

The sparklines show each product's trend in isolation. A `ggplot2` heatmap of
monthly sales across all products puts every product on the same colour scale,
so seasonal peaks and the strongest products stand out in a single grid. This is
a `ggplot`-only counterpart to the table-based views above.

```{r}
#| fig-height: 6
heat <- coffeechain %>%
  filter(year(Date) == 2013) %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales), .groups = "drop")

ggplot(heat, aes(x = Month,
                 y = fct_reorder(Product, Sales, .fun = sum),
                 fill = Sales)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c",
                      labels = scales::comma) +
  labs(title = "Monthly sales heatmap by product (2013)",
       x = NULL, y = NULL, fill = "Sales") +
  theme_few(base_size = 13) +
  theme(panel.grid = element_blank()) +
  black_text
```

::: callout-tip
## Insight
The heatmap trades the per-product detail of a sparkline for cross-product
comparison: every cell shares one colour scale, so the busiest products and the
seasonal pattern across the whole range are both readable at a glance.
:::

## 6.4 Trend small multiples by product type

The sparklines are intentionally minimal. When I want to read actual values, a
faceted area chart of monthly profit by `Product Type`, themed with `ggthemes`,
gives the same trend comparison with axes attached.

```{r}
#| fig-height: 5
type_trend <- coffeechain %>%
  filter(year(Date) == 2013) %>%
  mutate(Month = month(Date)) %>%
  group_by(`Product Type`, Month) %>%
  summarise(Profit = sum(Profit), .groups = "drop")

ggplot(type_trend, aes(Month, Profit)) +
  geom_area(fill = "#2166ac", alpha = 0.3) +
  geom_line(colour = "#2166ac", linewidth = 0.6) +
  facet_wrap(~ `Product Type`) +
  scale_x_continuous(breaks = seq(2, 12, 2)) +
  labs(title = "Monthly profit by product type (2013)",
       x = "Month", y = "Profit") +
  theme_few(base_size = 13) +
  black_text
```

## 6.5 Exception reporting: products below budget

Dashboards should flag problems, not just display data. This `gt` view keeps only
the products whose actual sales fell short of budget and colours the shortfall,
turning the table into a management-by-exception list.

```{r}
exceptions <- coffeechain %>%
  group_by(Product) %>%
  summarise(Actual = sum(Sales),
            Target = sum(`Budget Sales`)) %>%
  mutate(`Gap to target` = Actual - Target,
         `% of target` = Actual / Target) %>%
  filter(Actual < Target) %>%
  arrange(`% of target`)

exceptions %>%
  gt() %>%
  tab_header(title = "Products below budget",
             subtitle = "Management-by-exception view") %>%
  fmt_currency(columns = c(Actual, Target, `Gap to target`),
               currency = "USD", decimals = 0) %>%
  fmt_percent(columns = `% of target`, decimals = 1) %>%
  data_color(columns = `% of target`,
             palette = c("#b2182b", "#fddbc7")) %>%
  gt_theme_538()
```

::: callout-tip
## Insight
Filtering to the exceptions is itself a design decision. A reader scanning this
table sees only what needs action, which is the whole point of a dashboard
versus a raw data dump.
:::

# 7 Reflections

-   The strongest lesson is that **the right tool collapses effort**. The
    sparkline took roughly twenty lines in `ggplot2` (Section 3.2) and one line
    in `gtExtras` (Section 4.2). When a package is built for exactly one job, it
    is worth reaching for.
-   `gt`/`gtExtras` and `reactable`/`reactablefmtr` split cleanly by output:
    static and print-ready versus interactive and exploratory. Extras 1 and 3
    build the same product view in both, which made the trade-off concrete.
-   Dashboard design is mostly **editing**, not drawing. The exception view
    (Extra 5) and the colour-as-ranking scorecard (Extra 1) communicate faster
    than fuller tables precisely because they show less.
-   A practical snag worth noting: `gt_plt_bullet()` needs `svglite`, and the
    interactive sparklines need the GitHub-only `dataui` package. Getting the
    environment right was a larger share of the work than writing the plots.
